---
title: 'CSCI E-106:Assignment 2'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Brian Ledger

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggeffects)
library(ggplot2)
```

### Due Date: February 13, 2026 at 11:59 pm EST


## Problem 1 

_Refer to the CDI data set.  Use the number of active physicians as $Y$ and 
total personal income as $X$._

```{r}
cdi <- read.csv("CDI Data.csv")
cdi = cdi[,c("Total.personal.income", "Number.of.active.physicians")]
colnames(cdi) = c("TPI", "Active.Physicians")
```

_a-) Build a regression model to predict Y by using X. Write down the standard 
error of the slope and intercept._

```{r}
fit = lm(Active.Physicians ~ TPI, cdi)
summary(fit)
```

The standard error of the slope is 0.00211. 

The standard error of the intercept is 31.83333.

_b-) Are the slope and intercept statistically significant?  Write down your null and alternative hypothesis and p value of the test. Calculate the 95% Confidence interval for the slope._

The null hypothesis for the intercept is that the intercept is 0. The 
alternative hypothesis for the intercept is that the intercept is not 0.

The t-statistic for the intercept at -48.39485 is -1.52. The probability of
getting a t-statistic at least as large as -1.52 under the null hypothesis is 
0.129.

If I take a significance level to be 0.05, I cannot discharge the null
hypothesis, because 0.129 > 0.05.

--

The null hypothesis for the slope is that the slope is 0. The alternative 
hypothesis for the slope is that the slope is not 0.

The t-statistic for the slope at 0.13170 is 62.41. The probability of getting a
t-statistic at least as large as 62.41 is $2*10^{-16}$.

If I take a significance level to be 0.05, I can discharge the null
hypothesis, because $2*10^{-16}$ < 0.05.

--

The formula for the confidence interval of the coefficient of a linear
regression is:

$$
CI = \beta \pm t^*SE_{\beta} 
$$
where $\beta$ is the coefficient, $t^*$ is the critical value, and $SE_\beta$ is
the standard error of the coefficient.

In this case, $\beta$ is 0.13170, the critical value is 62.41, and the standard 
error is 0.00211.

```{r}
coefficients = summary(fit)$coefficients["TPI",]

estimate = coefficients[["Estimate"]]
standard_error = coefficients[["Std. Error"]]
degrees_of_freedom = df.residual(fit)

critical_value = qt(.095/2., degrees_of_freedom)

margin_of_error = critical_value * standard_error

lower_bound = estimate - margin_of_error
upper_bound = estimate + margin_of_error
```

The confidence interval for the slope is (`r lower_bound`, `r upper_bound`).

_c-) Predict the number of active physicians for total personal income of 
100,000, 250,000 and 50,000. Calculate the 95% confidence interval (Hint: Use 
the range of total personal income to identify which observations are within the 
range of the training data)_

```{r}
predictions = data.frame(TPI=c(100000, 250000, 50000))
predictions$Active.Physicians = predict(
  fit, newdata = predictions, interval="confidence", level=0.95
)
print(predictions)
```

```{r}
pred <- ggpredict(fit, terms = "TPI", ci.lvl = 0.95)
plot(pred)
```

The 95% confidence interval for active physicians where TPI = \$100,000 is 
$(12735.899, 13507.549)$.

The 95% confidence interval for active physicians where TPI = \$250,000 is 
$(31871.240, 33882.565)$.

The 95% confidence interval for active physicians where TPI = \$50,000 is 
$(6353.955, 6719.374)$.

_d-) What is the error variance of the  regression line?_

The formula for error variance of a regression line is

$$
S^2 = \frac{SSE}{n-2}
$$

The SSE and (n-2) are 569.7 and 438 respectively. This gives $S^2 = 1.300$.



## Problem 2 

Refer to the CDI data set. Use the number of active physicians as Y and total personal income as X. Select 1,000 random samples of 400 observations, fit the regression model and record $\beta_0$ and $\beta_1$ for each selected sample. Calculate the mean and variance of $\beta_0$ and $\beta_1$ based on the 1,000 different regression line and compare against the regression model in question 1 part b. (50 points)

```{r}
data = data.frame()
for (i in 1:1000) {
  sample = cdi[sample(nrow(cdi), 400), ]
  fit = lm(Active.Physicians ~ TPI, sample)
  data = rbind(data, coef(fit))
}
stats = data.frame(
  mean=c(mean(data[,1]), mean(data[,2])), 
  var=c(var(data[,1]), var(data[,2])), 
  stderr=c(sd(data[,1]), sd(data[,2]))
)
rownames(stats) = c("(Intercept)", "TPI")
print(stats)
```

The estimates of mean are pretty much the same. The variances/stderrs of the
intercepts differ by a factor of ~5.

The difference in variances of the intercepts may be less important because the 
intercept already fails to differ significantly from 0.

