---
title: 'CSCI E-106:Assignment 2'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Due Date: February 13, 2026 at 11:59 pm EST


## Problem 1 

Refer to the CDI data set.  Use the number of active physicians as Y and total personal income as X.

```{r}
cdi <- read.csv("CDI Data.csv")
summary(cdi)
```

```{r}
variables = names(cdi)
numeric_variables = names(cdi[, sapply(cdi, is.numeric)])
response_variable = "Number.of.active.physicians"
explanatory_variable = "Total.personal.income"
```

a-) Built a regression model to predict Y by using X. Write down the stand error of the slope and intercept.
```{r}
response_variable = "Number.of.active.physicians"
explanatory_variable = "Total.personal.income"

fit = lm(Number.of.active.physicians ~ Total.personal.income, cdi)
summary(fit)
```

b-) Are the slope and intercept statistically significant?  Write down your null and alternative hypothesis and p value of the test. Calculate the 95% Confidence interval for the slope. (20 points)

c-) Predict the number of active physicians for total personal income of 100000,250000 and 50000. Calculate the 95% confidence interval (Hint: Use the range of total personel income to identify which observations are within the range of the training data) (20 points)

d-) What is the error variance of the  regression line?(5 points)

## Problem 2 

Refer to the CDI data set. Use the number of active physicians as Y and total personal income as X. Select 1,000 random samples of 400 observations, fit the regression model and record $\beta_0$ and $\beta_1$ for each selected sample. Calculate the mean and variance of $\beta_0$ and $\beta_1$ based on the 1,000 different regression line and compare against the regression model in question 1 part b. (50 points)