---
title: 'CSCI E-106:Assignment 7_solutions'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## Helper to display regression function with n coefficients by Ian Kelk
## If var.names = TRUE add a space and "*" symbol to make readable
dispRegFunc <- function(reg, var.names=FALSE, pdfWrap=0, signif=6) {

  ## Set the line wrapping only if this is being knit as a pdf
  line.limit <- 0
  if (!is.null(knitr::current_input())) {
    if("pdf_document" %in% rmarkdown::all_output_formats(knitr::current_input())) {
       line.limit <- pdfWrap
    }
  }
  ## Check to see if this is a cv.glmnet object or a lm or rlm object
  ifelse(class(reg) == "cv.glmnet",
         coefs <- predict(reg, s=reg$lambda.min, type="coefficients")[,1],
         coefs <- reg$coefficients)
  
  b0 = coefs[1]
  n <- length(coefs)
  my_formula <- paste0("Y = ", round(b0, digits = signif))
  for (i in 2:n) {
    ## Choose the +/- sign for the coefficient then remove it using abs()
    options(scipen = 999)
    new.formula.term <- paste0(ifelse(coefs[i] < 0," - "," + "),
                         signif(abs(coefs[i]),signif),
                         ifelse(var.names,paste0(" * ",names(coefs)[i]),
                         names(coefs)[i]))
    if (line.limit > 0) {
      if (line.limit <= nchar(my_formula) + nchar(new.formula.term)) {
        print(my_formula)
        my_formula <- new.formula.term
      } else {
        my_formula <- paste0(my_formula, new.formula.term)
      }
    }
    else {
      my_formula <- paste0(my_formula, new.formula.term)
    }
  }
  print(my_formula)
}
# Function to load or download libraries
loadLib = function(libName)
{
    if(require(libName, character.only=TRUE))
    {
        ##cat(libName, "loaded properly\n")
    } else {
        cat("Installing", libName, "\n")
        install.packages(libName)
        if(require(libName, character.only=TRUE))
        {
            cat(libName, "loaded properly\n")
        } else {
            stop(c(libName, "not properly installed\n"))
        }
    }
}


```

### Due Date: March 27, 2026 at 11:59 pm EST
### Instructions

Students should submit their reports on Canvas. The report needs to clearly state what question is being solved, step-by-step walk-through solutions, and final answers clearly indicated. Please solve by hand where appropriate.

Please submit two files: (1) a R Markdown file (.Rmd extension) and (2) a PDF document, word, or html generated using knitr for the .Rmd file submitted in (1) where appropriate. Please, use RStudio Cloud for your solutions.

--------------------------------------------------------------------------------------

## Problem 1

Using the fat data under faraway library,if you are using the R Studio, please use the command below to download the data:
```{r}
library(faraway)
data(fat,package="faraway")
#str(fat)
```

If you are using Posit R Cloud instead of RStudio Desktop, please use the command below to download the data:

```{r}
tmp <- tempfile(fileext = ".rda")
download.file(
  "https://raw.githubusercontent.com/cran/faraway/master/data/fat.rda",
  destfile = tmp,
  mode = "wb"
)
load(tmp)
head(fat)
```
Take the fat data, and use the percentage of body fat, siri, as the response and the other variables, except brozek and density as potential predictors. (75 points) 

Data Exploration, please note that the code below remove brozek and density as they must be excluded as potential predictors.

```{r,include=TRUE, message=FALSE, warning=FALSE}
#load data
head(fat) #need to exclude brozek and density, that is [, -c(1,3)]
fat=fat[, -c(1,3)]
head(fat) #verify that brozek and density are not there
summary(fat) 
#check for missing and NAs
apply(fat ,2,function(x) sum(is.na(x)))
#Are there any candidate for dummies?
### No
```
Use the cor() R function rounded to 2 digits to obtain the correlation matrix and the plot() R function to obtain the scatter plot

```{r,message=FALSE, warning=FALSE}
Correlations<-round(cor(fat),1)
Correlations
#Findings
#several high correlated predictors (above 0.6): hip and weight, and weight against neck, chest, abdom, hip or thigh

```


## a1-) Create train and test data sets. Use 70% of the data for train data set and use remaining data (30% of data) for test data set (use set.seed(1023)). (5 points)
**Conclusion** The train data size is close to 70% (it is 0.6984127) so it is OK.
Data Preparations
```{r, include=FALSE}
#randomizing and partitioning the data
 #in order to get same partitioning every time
n <- dim(fat)[1]
set.seed(1023)
ind <- sample(1:n, n*0.7)
data_train <- data.frame(fat[ind, ])
data_test <- data.frame(fat[-ind, ])
#always check the dimensions in each dataset (error checking) 
dim(fat)
dim(data_train)
dim(data_test) 
176/252  #
```

**Use the train data sets to answer questions below**

## a2-)Linear regression with all predictors and comment on the model performance (10 points)
**Conclusion** The regression model is "Y = -65.175022 + 0.0165101age + 0.258471weight + 0.863336height + 0.630104adipos - 0.596346free + 0.0361831neck + 0.103261chest + 0.0747285abdom - 0.0378032hip + 0.196779thigh + 0.088616knee + 0.0948155ankle + 0.0667037biceps + 0.253269forearm + 0.263196wrist". Normal linear regression model does not seem to be appropriate for the problem. The R^2 is extremely high and there were 9 not significant variables. 
<!-- Input solution below -->
```{r}
lmod <- lm(siri~.,data=data_train)
summary(lmod)
dispRegFunc(lmod)
```
<!-- End of solution -->

## b-)Linear regression with variables selected using stepwise (both ways) selection criteria  (10 points)
**Conclusion** The linear model using stepwise (both ways) selection criteria and all predictors are statistically significant at an alpha of 0.05 is "Y = -61.215606 - 0.601572free + 0.281504weight + 0.323059forearm + 0.901129height + 0.118269chest + 0.210532thigh + 0.035185age + 0.705783adipos". Although more analysis is needed due to the presence of outliers.
<!-- Input solution below -->
```{r}
library(olsrr)
# Forward-Backward Stepwise Regression by p-value
bothlm = ols_step_both_p(lmod, pent=0.05, prem=0.05)
summary(bothlm$model)
# wrist is not signiificative so we drop it from the final selection
bothlmod <-lm(siri~free+weight+forearm+height+chest+thigh+age+adipos,data=data_train)
summary(bothlmod)
dispRegFunc(bothlmod)
par(mfrow=c(2,2))
plot(bothlmod)

```
<!-- End of solution -->

## c-)Perform Ridge Regression and comment on the model performance (10 points)
**Conclusion** The Ridge regression model with the optimal lambda and coefficients for our data is:"Y= -68.28916029 + 0.04832933age + 0.05713141weight + 0.49130278height" 
See code and explantations below:
<!-- Input solution below -->
```{r, message=FALSE, warning=FALSE}
library(glmnet)
#need to define X and Y matrix for glmnet
#take out Y and law indicator from data for X (all independent variable)
x <- model.matrix(siri~.,data=data_train)[,-c(1)]
y <- data_train$siri
RidgeMod <- glmnet(x, y, alpha=0, nlambda=100,lambda.min.ratio=0.0001)
```
In the glmnet function are used the following arguments:

-nlambda=100: Set the number of lambda values (the default is 100)

-lambda.min.ratio=0.0001: Set the smallest value for lambda, as a fraction of lambda.max, the (data derived) entry value (that is, the smallest value for which all coefficients are zero)

we will use cross valdidation to select the best lamda

```{r}
library(glmnet)
CvRidgeMod <- cv.glmnet(x, y, alpha=0, nlambda=100,lambda.min.ratio=0.0001)
par(mfrow=c(1,1))
plot(CvRidgeMod)
best.lambda.ridge <- CvRidgeMod$lambda.min
best.lambda.ridge 
```

The figure includes the cross-validation curve (red dotted line) and upper and lower standard deviation curves along the $\lambda$ sequence (error bars). In the beginning of the procedure (to the right of the figure), the MSE is very high, and the coefficients are restricted to be too small; and then at some point, it kind of levels off. This seems to indicate that the full model is doing a good job.

There are two vertical lines: one is at the minimum, and the other vertical line is within one standard error of the minimum. The second line is a slightly more restricted model that does almost as well as the minimum, and sometimes, we'll go for that.
These lines then lie at two lambda values:

-lambda.min is the value of $\lambda$ that gives the minimum mean cross-validated error

-lambda.1se, gives the most regularized model such that the error is within one standard error of the minimum

At the top of the plot, you actually see how many nonzero variables' coefficients are in the model. There are all six variables in the model (five variables, plus the intercept), and no coefficient is zero.

<!-- End of solution -->
Once we have the best lambda, we can use predict to obtain coefficients:This is the best regression model for our data.
```{r}
predict(RidgeMod, s=best.lambda.ridge, type="coefficients")[1:4, ]


```

## d-)Perform Lasso Regression and comment on the model performance (10 points)
**Conclusion** The Lasso regression model with the optimal lambda and coefficients for our data is: "Y= -53.18757603 + 0.01599433age + 0.28264029weight + 0.65254778height + 0.32722328adipos - 0.58736917free + 0.03139386neck + 0.1124703chest + 0.08822504abdom + 0.17806886thigh + 0.09217264knee + 0.09818182ankle
+ 0.08013646biceps + 0.26251115forearm + 0.22830567wrist . The Lasso regression model for our data only eliminates one variable see below, in order to perform Lasso regression, we used the R function below and $\alpha = 1$ for Lasso regression. Everything else is very similar to the Ridge regression.
<!-- Input solution below -->
```{r}
set.seed(1023)   #set seed here to ensure replicability of the results
library(glmnet)
x <- model.matrix(siri~.,data=data_train)[,-c(1)]
y <- data_train$siri
LassoMod <- glmnet(x, y, alpha=1, nlambda=100,lambda.min.ratio=0.0001)
plot(LassoMod,xvar="norm",label=TRUE)
CvLassoModel <- cv.glmnet(x, y, alpha=1, nlambda=100,lambda.min.ratio=0.0001)
plot(CvLassoModel)
best.lambda.lasso <- CvLassoModel$lambda.min
best.lambda.lasso
coef(CvLassoModel, s = "lambda.min")
```
<!-- End of solution -->
We have confirmed that the Lasso method is able to make a selection of variables. Ultimately, we can say that both Lasso and Ridge balance the trade-off bias-variance with the choice of $\lambda$. Lasso implicitly assumes that part of the coefficients are zero, or at least not significant. Lasso tends to have a higher performance than Ridge in cases where many predictors are not actually tied to the response variables. In opposite cases, the Ridge tends to have better performance. Both approaches can be compared by cross-validation.


## e-)Perform Elastic Net and comment on the model performance (10 points)
**Conclusion** The Elastic Net regression model with the optimal lambda and coefficients for our data is: "Y= -73.38371206 + 0.01539566age + 0.20285358weight + 0.89562777height + 0.69359127adipos - 0.56182464free +
0.04217094neck + 0.10868605chest + 0.12922567abdom + 0.18864365thigh + 0.08733581knee + 0.10142691ankle + 0.09580345biceps + 0.26522297forearm + 0.20031193wrist.  The Elastic Net regression model for our data is different from Lasso as the coefficients changed. Only hip is eliminated.
See code and explanations below:
<!-- Input solution below -->
```{r}
set.seed(1023)   #set seed here to ensure replicability of the results
EnetMod <- glmnet(x, y, alpha=0.5, nlambda=100,lambda.min.ratio=0.0001)
CvElasticnetMod <- cv.glmnet(x, y,alpha=0.5,nlambda=100,lambda.min.ratio=0.0001)
best.lambda.enet <- CvElasticnetMod$lambda.min
best.lambda.enet
coefficients(EnetMod,s=best.lambda.enet)
#coef(CvElasticnetMod, s = "lambda.min")
```
<!-- End of solution -->

**Use the test data sets to answer questions below**

## f-) Test the model's performances  for part a to part e above and comment of model stability and performance. Which model would you choice and why? (20 points)

**Conclusion** Based on performance on test data Ridge Regression performed better when using Rsquare or RMSE, however the Lasso regression is close and it has the lowest MAE. I would choose the Ridge regression because it has the minimum RMSE and largest Rsquared.
<!-- Input solution below -->

```{r, message=FALSE, warning=FALSE}
library(caret)
#performance of the model on the Test Data
#getting the predicted values for the test data linear model
PredictedTest<-predict(lmod, data_test)
ModelTest<-data.frame(obs = data_test$siri, pred=PredictedTest)
#getting the predicted values for the test data both ways model selection 
PredictedTest2<-predict(bothlmod, data_test)
ModelTest2<-data.frame(obs = data_test$siri, pred=PredictedTest2)
#getting the predicted values for the test data Ridge model 
newx<-model.matrix(siri~., data_test)[,-c(1)]
PredictedTest3 <- predict(CvRidgeMod, s = best.lambda.ridge, newx = newx)
ModelTest3<-data.frame(obs = data_test$siri, pred=PredictedTest3)
# need a little trick here
colnames(ModelTest3)[2]="pred"
#getting the predicted values for the test data Lasso model
PredictedTest4 <- predict(CvLassoModel, s = best.lambda.lasso,newx = newx)
ModelTest4<-data.frame(obs = data_test$siri, pred=PredictedTest4)
colnames(ModelTest4)[2]="pred"
#getting the predicted values for the test data Elastic net model
PredictedTest5 <- predict(CvElasticnetMod, s = best.lambda.enet,newx = newx)
ModelTest5<-data.frame(obs = data_test$siri, pred=PredictedTest5)
colnames(ModelTest5)[2]="pred"

#comparing the model performance metrics in a small report

out<-rbind(defaultSummary(ModelTest),defaultSummary(ModelTest2),defaultSummary(ModelTest3),defaultSummary(ModelTest4),defaultSummary(ModelTest5))
dimnames(out)[[1]]<-c("Linear","Both Ways S", "Ridge","Lasso","Elastic")
out
```
<!-- End of solution -->


## Problem 2

Use the fat data set in previous example by using the percentage of body fat, siri, as the response and the other variables, except brozek and density as potential predictors.(25 points)
Data Preparations
```{r, include=FALSE}

#randomizing and partitioning the data
head(fat) #verify that brozek and density are not there so we can use the problem 1 training and test data
```

## a-)Fit the same model in question 1-a but now using Huber's robust method. Comment on any substantial differences between this model and the least squares fit. (10 points)

**Conclusion** Here we use the train data for an apples to apples comparison with the model in question 1-a. When establishing a rule of thumb of 60% variation as not good, the predictors named: forearm, knee, hip and chest show the largest variation of coefficients revealing the presence of multiple influential outlier points. Here we use the 1-a model training dataset for the regressions and the Huber option psi=psi.huber for the huber robust method as prescribed in the question. See code below.
<!-- Input solution below -->
```{r, message=FALSE, warning=FALSE}
set.seed(1023)  #to ensure replicability of the results
library(MASS)
lmod2<-lm(siri~., psi = psi.huber, data=data_train)  # same model as before
rr.huber2 <- rlm(siri~., psi = psi.huber, data=data_train)
summary(rr.huber2)

rr.summary <- summary(rr.huber2)
lmod.summary <- summary(lmod2)

percentage_difference<-round(((lmod2$coefficients-rr.huber2$coefficients)/lmod2$coefficients)*100,0)

coef_df <- data.frame(
           "Linear Model" = lmod2$coefficients, 
           "RobustHuber"= rr.huber2$coefficients,
           "Percentage Difference"=percentage_difference
            )

dispRegFunc(lmod2)
dispRegFunc(rr.huber2)

knitr::kable(coef_df, caption="Co-efficients Comparison")

par(mfrow=c(2,2))
plot(lmod2)
plot(rr.huber2)
```
<!-- End of solution -->



## b-)Identify which two cases have the lowest weights in the Huber fit. What is unusual about these two points? (5 points)
**Conclusion**Here you need to remember that the Huber method works multiplying the empirical residuals by a factor less than one, the output is named weighted residuals and can be extract for analysis using the code below. The are ordered from the lowest ("weight") to the highes so if you review the first five you can answer the question. The observations 39 and 169 may be influential outliers as we note they have the lowest weights in the Huber fit. This result is expected as the method try to mitigate the impact of extreme cases by ponderating the weighting through the middle observations.
<!-- Input solution below -->
```{r}
#rr.huber2$w
hweights <- data.frame( resid = rr.huber2$resid, weight = rr.huber2$w)
hweights1 <- hweights[order(rr.huber2$w), ]
hweights1[1:5, ]
```
<!-- End of solution -->

## c-)Plot weight (of the man) against height. Identify the two outlying cases. Are these the same as those identified in the previous question? Discuss.(10 points)
**Conclusion** The graph shows other outlier points detected in the Huber regression in red ink. We note graph shows other outliers like point 28 or point 34 at the top of the figure. From our inspection of the plot we conclude that these points are not the same as the ones in the Huber regression as outlier points in the direction of the regression may not be influential in the coefficient determination.
<!-- Input solution below -->
```{r}
plot(
  data_train$weight,
  data_train$height,
  xlab = "Weight",
  ylab = "Height",
  main = "Weight Vs Height"
)

# plot all labels in black
text(
  data_train$weight,
  data_train$height,
  labels = 1:176,
  pos = 3,
  cex = 0.7
)

# highlight observations 39 and 169 in red
points(
  data_train$weight[c(39, 169)],
  data_train$height[c(39, 169)],
  col = "red",
  pch = 19
)

text(
  data_train$weight[c(39, 169)],
  data_train$height[c(39, 169)],
  labels = c(39, 169),
  col = "red",
  pos = 3,
  cex = 0.8
)
```




<!-- End of solution -->

