---
title: "Regression Tree"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Fitting Regression Trees 

Here we fit a regression tree to the `Boston`  data set. First, we create a training set, and fit the tree to the training data.

```{r chunk14}
Boston <- read.csv("/cloud/project/Data/Boston.csv")
head(Boston)
#there is an index variable named X, dropping it from the data set
Boston <- Boston[-1]
library(tree)
set.seed(1)
#50 to 50 split
train <- sample(1:nrow(Boston), nrow(Boston) / 2)
tree.boston <- tree(medv ~ ., Boston, subset = train)
summary(tree.boston)
```

Notice that the output of `summary()` indicates that only four of the variables have been used in constructing the tree. In the context of a regression tree, the deviance is simply the sum of squared errors for the tree. We now plot the tree.

```{r chunk15}
plot(tree.boston)
text(tree.boston, pretty = 0)
```

The variable `lstat` measures the percentage of individuals with {lower  socioeconomic status}, while the variable `rm` corresponds to the average number of rooms. The tree indicates that larger values of `rm`, or lower values of `lstat`, correspond to more expensive houses. For example, the tree predicts a median house price of $$45{,}400$ for homes in census tracts in which `rm >= 7.553`.



```{r}
library(caret)
#performance of the model on the TrainData
#get the predicted values
PredictedTrain<- predict(tree.boston, newdata = Boston[train, ])
#create a data frame with 2 columns actual and predicted values
ModelTrain<-data.frame(obs = Boston[train, ]$medv, pred=PredictedTrain)
#calculate performance statistics
tree1<-defaultSummary(ModelTrain)

#lets compare this against regression model
f<-lm(medv~.,data=Boston[train,])
summary(f)
PredictedTrain.reg<- predict(f, newdata = Boston[train, ])
#create a data frame with 2 columns actual and predicted values
ModelTrain.reg<-data.frame(obs = Boston[train, ]$medv, pred=PredictedTrain.reg)
#calculate performance statistics
reg1<-defaultSummary(ModelTrain.reg)
out<-rbind(tree1,reg1)
out
```

Regression tree performs better than OLS regression model on the train data set.


It is worth noting that we could have fit a much bigger tree, by
passing `control = tree.control(nobs = length(train), mindev = 0)` into the `tree()` function.

Now we use the `cv.tree()` function to see whether pruning the tree will improve performance.

```{r chunk16}
cv.boston <- cv.tree(tree.boston)
plot(cv.boston$size, cv.boston$dev, type = "b")
```

In this case, the most complex tree  under consideration is selected by cross-validation. However, if we wish to prune the tree, we could do so as follows, using the `prune.tree()` function:

```{r chunk17}
prune.boston <- prune.tree(tree.boston, best = 5)
plot(prune.boston)
text(prune.boston, pretty = 0)
```

In keeping with the cross-validation results, we use the unpruned tree to make predictions on the test set.

```{r chunk18}
yhat <- predict(tree.boston, newdata = Boston[-train, ])
boston.test <- Boston[-train, "medv"]
plot(yhat, boston.test)
abline(0, 1)
mean((yhat - boston.test)^2)


#performance of the model on the Test data
#get the predicted values
PredictedTest<- predict(tree.boston, newdata = Boston[-train, ])
#create a data frame with 2 columns actual and predicted values
ModelTest<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTest)
#calculate performance statistics
tree1.test<-defaultSummary(ModelTest)

#lets compare this against regression model

PredictedTrain.reg.test<- predict(f, newdata = Boston[-train, ])
#create a data frame with 2 columns actual and predicted values
ModelTrain.reg.test<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTrain.reg.test)
#calculate performance statistics
reg1.test<-defaultSummary(ModelTrain.reg.test)
out.test<-rbind(tree1.test,reg1.test)
out.test
```
Again, regression tree outperforms OLS regression on the test data set.

In other words, the  test set MSE associated with the regression tree is $35.29$.
The square root of the MSE is therefore around $5.941$, indicating that this model leads to test predictions that are (on average) within approximately $$5{,}941$ of the true median home value for the census tract.

## Bagging and Random Forests

Here we apply bagging and random forests to the `Boston` data, using the `randomForest` package in `R`. The exact results obtained in this section may depend on the version of `R` and the version of the `randomForest` package installed on your computer.
Recall that bagging is simply a special case of a random forest with
$m=p$. Therefore, the
`randomForest()` function can be used to perform both random forests and bagging.
We perform bagging as follows:


```{r chunk19}
library(randomForest)
Boston <- read.csv("/cloud/project/Data/Boston.csv")
set.seed(1)
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, importance = TRUE)
bag.boston
```

The argument `mtry = 12` indicates that all $12$ predictors should be considered for each split of the tree---in other words, that bagging should be done.
How well does this bagged model perform on the test set?

```{r chunk20}
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
PredictedTest.rf<- predict(bag.boston, newdata = Boston[-train, ])
#create a data frame with 2 columns actual and predicted values
ModelTest.rf<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTest.rf)
#calculate performance statistics
rf.test<-defaultSummary(ModelTest.rf)
out.test1<-rbind(out.test,rf.test)
out.test1
plot(yhat.bag, boston.test)
abline(0, 1)
mean((yhat.bag - boston.test)^2)
```

The test set MSE associated with the bagged regression tree is $23.42$, about two-thirds of that obtained using an optimally-pruned single tree.
We could change the number of trees grown by `randomForest()` using the `ntree` argument:

```{r chunk21}
bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, ntree = 25)
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
mean((yhat.bag - boston.test)^2)

PredictedTest.rf<- predict(bag.boston, newdata = Boston[-train, ])
#create a data frame with 2 columns actual and predicted values
ModelTest.rf<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTest.rf)
#calculate performance statistics
defaultSummary(ModelTest.rf)
```


Growing a random forest proceeds in exactly the same way, except that we use a smaller value of the `mtry` argument. By default, `randomForest()` uses $p/3$ variables when building a random forest of regression trees, and $\sqrt{p}$ variables when building a random forest of classification trees. Here we use `mtry = 6`.

```{r chunk22}
set.seed(1)
rf.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 6, importance = TRUE)
yhat.rf <- predict(rf.boston, newdata = Boston[-train, ])
mean((yhat.rf - boston.test)^2)

bag.boston <- randomForest(medv ~ ., data = Boston,
    subset = train, mtry = 12, ntree = 25)
yhat.bag <- predict(bag.boston, newdata = Boston[-train, ])
mean((yhat.bag - boston.test)^2)

PredictedTest.rf<- predict(rf.boston, newdata = Boston[-train, ])
#create a data frame with 2 columns actual and predicted values
ModelTest.rf<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTest.rf)
#calculate performance statistics
defaultSummary(ModelTest.rf)
```

The test set MSE is $20.07$; this indicates that random forests yielded an improvement over bagging in this case.

Using the `importance()` function, we can view the importance of each variable.

```{r chunk23}
importance(rf.boston)
```

Two measures of variable importance are reported. The first is based upon the mean decrease of accuracy in predictions on the out of bag samples when a given variable is permuted. The second is a measure of the total decrease in node impurity that results from splits over that variable, averaged over all trees (this was plotted in Figure 8.9). In the case of regression trees, the node impurity is measured by the training RSS, and for classification trees by the deviance. Plots of these importance measures can be produced using the `varImpPlot()` function.

```{r chunk24}
varImpPlot(rf.boston)
```

The results indicate that across all of the trees considered in the random forest, the wealth of the community (`lstat`) and the house size (`rm`) are by far the two most important variables.

## Boosting


Here we use the `gbm` package, and within it the `gbm()` function, to fit boosted regression trees to the `Boston` data set. We run `gbm()` with the option `distribution = "gaussian"` since this is a regression problem; if it were a binary classification problem, we would use `distribution = "bernoulli"`.
The argument `n.trees = 5000` indicates that we want $5000$ trees, and the option `interaction.depth = 4` limits the depth of each tree.

```{r chunk25}
library(gbm)
set.seed(1)
boost.boston <- gbm(medv ~ ., data = Boston[train, ],distribution = "gaussian", n.trees = 5000,interaction.depth = 4)
```

The `summary()` function produces a relative influence plot and also outputs the relative influence statistics.

```{r chunk26}
summary(boost.boston)
```

We see that `lstat` and `rm` are by far the most important variables. We can also produce *partial dependence plots* for these two variables. These plots illustrate the marginal effect of the selected variables on the response after *integrating* out the other variables. In this case, as we might expect, median house prices are increasing with `rm` and decreasing with `lstat`.

```{r chunk27}
plot(boost.boston, i = "rm")
plot(boost.boston, i = "lstat")
```


We now use the boosted model to predict `medv` on the test set:

```{r chunk28}
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)

PredictedTest.gbm<- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
#create a data frame with 2 columns actual and predicted values
ModelTest.gbm<-data.frame(obs = Boston[-train, ]$medv, pred=PredictedTest.gbm)
#calculate performance statistics
defaultSummary(ModelTest.gbm)

```

The test MSE obtained is $18.39$: this is superior to the test MSE of random forests and bagging. If we want to, we can perform boosting with a different value of the shrinkage parameter $\lambda$ in (8.10). The default value is $0.001$, but this is easily modified.
Here we take $\lambda=0.2$.

```{r chunk29}
boost.boston <- gbm(medv ~ ., data = Boston[train, ],
    distribution = "gaussian", n.trees = 5000,
    interaction.depth = 4, shrinkage = 0.2, verbose = F)
yhat.boost <- predict(boost.boston,
    newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```

In this case, using $\lambda=0.2$ leads to a lower test MSE than $\lambda=0.001$.

###using RPART Library

```{r}
wine<- read.csv("/cloud/project/Data/whitewines.csv")
str(wine)
head(wine ,10)
hist(wine$quality)
```

The wine quality values appear to follow a fairly normal, bell-shaped distribution, centered around a value of six. This makes sense intuitively, because most wines are of average quality; few are particularly bad or good. 


```{r}
summary(wine)
```

There are outliers in the data. Even though trees are fairly robust to messy data, it is always prudent to check for severe problems. For now, we'll assume that the data is reliable.

lets do 70 and 30 split


```{r}
set.seed(1023)
sample_ind<-sample(nrow(wine),round(0.7*nrow(wine)))
wine_train <- wine[sample_ind, ]
wine_test <- wine[-sample_ind, ]
#install.packages("rpart") 
library(rpart)
m.rpart <- rpart(quality ~ ., data = wine_train)
m.rpart

```


For each node in the tree, the number of examples reaching the decision point is listed. For instance, all 3,429 examples begin at the root node, of which 2,155 have alcohol < 10.85 and 1,274 have alcohol >= 10.85. Because alcohol was used first in the tree, it is the single most important predictor of wine quality.

Nodes indicated by * are terminal or leaf nodes, which means that they result in a prediction (listed here as yval). For example, node 4 has a yval of 5.367. When the tree is used for predictions, any wine samples with alcohol < 10.85 and volatile.acidity < 0.2525 would therefore be predicted to have a quality value of 5.367.


```{r}
library(rpart.plot)
rpart.plot(m.rpart, digits = 3)
```

The fallen.leaves parameter forces the leaf nodes to be aligned at the bottom of the plot, while the
type and extra parameters affect the way the decisions and nodes are labeled. 
The numbers 3 and 101 refer to specific style formats, which can be found in the command's documentation, 
or via experimentation with various numbers.


```{r}
rpart.plot(m.rpart, digits = 4, fallen.leaves = TRUE,type = 3, extra = 101)
#evaluating model performance
p.rpart <- predict(m.rpart, wine_test)
summary(p.rpart)
summary(wine_test$quality)
cor(p.rpart, wine_test$quality)


PredictedTest1<- predict(m.rpart, wine_test)
#create a data frame with 2 columns actual and predicted values
ModelTest1<-data.frame(obs = wine_test$quality, pred=PredictedTest1)
#calculate performance statistics
defaultSummary(ModelTest1)


```

A correlation of 0.53 is certainly acceptable. However, the correlation only measures how strongly the 
predictions are related to the true value; it is not a measure of how far off the predictions were from 
the true values.


```{r}
library(caret)
PredictedTest<-predict(m.rpart, wine_test)
ModelTest<-data.frame(obs = wine_test$quality, pred=PredictedTest)
defaultSummary(ModelTest)
```

##Pruning

cp: complexity parameter. Any split that does not decrease the overall lack of fit by a 
factor of cp is not attempted. For instance, with anova splitting, this means that the overall 
R-squared must increase by cp at each step. The main role of this parameter is to save computing 
time by pruning off splits that are obviously not worthwhile. Essentially,the user 
informs the program that any split which does not improve the fit by cp will likely be pruned off 
by cross-validation, and that hence the program need not pursue it.
```{r}
library(faraway)
set.seed(123)
tmod<-rpart(O3~.,ozone,cp=0.001)
printcp(tmod)
plotcp(tmod)
```

In this table, we see the value of the cp parameter, the number of splits in the tree, the RSS of the tree divided by the RSS of the null tree.\newline
xerror denotes the cross-validated error which is also scaled by the RSS of the null tree. Since the partition of the data into 10 parts is random, 
this CV error is also random, which makes the given standard error useful. The random division also means that if you repeat this command, 
you will not get the same answer unless you use set.seed. We can select the size of the tree by minimizing the value of xerror and selecting the corresponding value of CP.

We can select the smallest tree with a CV error within one standard error of the minimum — in this case, 0.350 + 0.035 = 0.385. 
So we would take the seven-split tree.

```{r}
library(rpart.plot)
rpart.plot(tmod,type=3)
```
