---
title: 'CSCI E-106:Assignment 4'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(lmtest))
library(MASS)
```

### Due Date: February 27, 2026 at 11:59 pm EST


## Problem 1

\emph{Refer to Sales growth Data.}

```{R}
set.seed(1023)
sales_growth_data <- read.csv("Sales Growth Data.csv")
data = sales_growth_data
```

a-) \emph{Divide the range of the predictor variable (coded years) into five 
bands of width 2.0, as follows: Band 1 ranges from X = -.5 to X = 1.5; band 2 
ranges from X = 1.5 to X = 3.5; and so on. Determine the median value of X and 
the median value of Y in each band and develop the band smooth by connecting the 
five pairs of medians by straight lines on a scatter plot of the data.}  

```{R}
getMedians <- function(data, width, startX) {
  endX = startX + width
  selection = data[data$X > startX & data$X < endX,]
  return(c(X=median(selection$X), Y=median(selection$Y)))
}

par(mfrow=c(1,1))
plot(data$X, data$Y)

medians = getMedians(data,2,-0.5)
for(i in 1:4) {
  nextMedians = getMedians(data,2,-0.5+2*i)
  segments(medians[["X"]], medians[["Y"]], nextMedians[["X"]], nextMedians[["Y"]])
  medians = nextMedians
}

```

\emph{Does the band smooth suggest that the regression relation is linear?
Discuss.}

The regresssion relation for this data is clearly linear. Generally, there are
examples of non-linear data that would have band smooth. For instance, high 
frequency changes in $Y$ will be smooth for some band widths and jagged for
others. Data with concentrated values can appear to have medians higher or lower
than the means, which can smooth or unsmooth the connections.

Data that is linear may tend in general to have band smooth.

b-) \emph{Create a series of seven overlapping neighborhoods of width 3.0 
beginning at X = -.5. The first neighborhood will range from X = -.5 to X = 2.5; 
the second neighborhood will range from X = .5 to X = 3.5; and so on. For each 
of the seven overlapping neighborhoods, fit a linear regression function and 
obtain the fitted value $\hat{Y_{c}}$ at the center $X_{c}$ of the neighborhood. 
Develop a simplified version of the lowess smooth by connecting the seven 
($X_c$, $\hat{Y_{c}}$) pairs by straight lines on a scatter plot of the data.}

```{R}
plotB = function(data) {
  getCenter <- function(data, width, startX) {
    midX = startX + (width/2)
    endX = startX + width
    selection = data[data$X > startX & data$X < endX,]
    
    selectionFit = lm(Y~X, selection)
    res = data.frame(X=c(midX))
    res$Y = predict(selectionFit, res)
    return(res)
  }
  
  plot(data$X, data$Y)
  
  center = getCenter(data, 3, -0.5)
  for(i in 1:6) {
    nextCenter = getCenter(data, 3, -0.5 + i)
    segments(center[["X"]], center[["Y"]], nextCenter[["X"]], nextCenter[["Y"]])
    center = nextCenter
  }
}
```

c-) \emph{Obtain the 95 percent confidence band for the true regression line and
plot it on the plot prepared in part}

```{R}
lm_ = lm(Y~X, data)
ci = data.frame(X=c(min(data$X), max(data$X)))
ci = cbind(ci, predict(lm_, ci, interval="confidence"))

plotB(data)
segments(ci[1,"X"], ci[1,"lwr"], ci[2,"X"], ci[2,"lwr"])
segments(ci[1,"X"], ci[1,"upr"], ci[2,"X"], ci[2,"upr"])
```

(b). \emph{Does the simplified lowess smooth fall entirely within the confidence 
band for the regression line?}

Yes.

\emph{What does this tell you about the appropriateness of the linear regression 
function?}

Because the simplified loess smooth is a closer fit to the data, it would tend
to exceed the confidence band for data that doesn't follow a linear regression.
That the simplified loess smooth falls inside of the confidence band of the
linear regression indicates that the linear regression matches the closer fit,
that the closer fit is close to linear, and therefore the linear regression is
a good fit for the data.

## Problem 2

Using the ozone data under faraway library,if you are using the R Studio, please use the command below to download the data:
```{R}
library(faraway)
data(ozone,package="faraway")
data = ozone
```

\emph{Use O3 as the response and temp, humidity and ibh as predictors to answer 
the questions below.}

```{R}
data = data[, c("O3", "temp", "humidity", "ibh")]
k = 3
```

a-) \emph{Create train and test data sets. Use 70% of the data for train data 
set and use remaining data (30% of data) for test data set (use 
set.seed(1023)).}

```{R}
popN = nrow(data)
alphaN =7/10*popN
idx_train = sample(popN, size = alphaN, replace=FALSE)
train = data[idx_train,]
trainN = nrow(train)
test = data[-idx_train,]
testN = nrow(test)
```

b-) \emph{Build a regression model to predict O3 on the train data set. Write 
down the regression model and comment on the model performance.}

```{R}
lm_ = lm(O3 ~ ., data)
lm_summary = summary(lm_)
print(lm_summary)
```

The P-values are all significantly less than 0.05.

c-) \emph{Visually by using the graphs, comments on regression model 
assumptions.}

```{R}
par(mfrow=c(1,2))
plot(lm_, which=c(1,3))
```
The residuals increase toward the higher extrema. The residuals are unevenly 
distributed at the lower extreme. 

```{R}
plot(lm_, which=2)
```
The Q-Q plot has deviation at the ends.

d-) \emph{Test the model performance on the test data and comment on model 
stability and performance.}

```{R}
predictTest <-predict(lm_, test)
modelTest <- cbind(predictTest, test)
testResiduals = modelTest$O3 - modelTest$predictTest
dimE = testN - k - 1

sst = sum((modelTest$O3-mean(modelTest$O3))^2)
sse = sum(testResiduals^2)
ssr = sum((modelTest$predictTest - mean(modelTest$O3))^2)

mse = sse/dimE
rse = sqrt(mse)
rmse = sqrt(sse/length(testResiduals))
rSquared = 1.0 - (sse/sst)
mae = sum(testResiduals)/length(testResiduals)

model_summary = data.frame(MSE=c(mse), RSE=rse, RMSE=c(rmse), 
                           Rsquared=c(rSquared), MAE=c(mae))
print(model_summary)
```

The $R^2$ in the training set was `r round(lm_summary$r.squared, 3)`. 
In the test it is `r round(model_summary$Rsquared, 3)`.

e-) \emph{Perform Breusch-Pagan Test, write down null and alternative hypotheses 
and your conclusion.}

$H_0$: The error variances in the regression residuals are homoscedastic.

$H_a$: The error variances in the regression residuals are heteroscedastic.

```{R}
bptest(lm_)
```

In the linear model, the p-value is small, so we accept the alternative 
hypothesis. The residual variances are heteroscedastic.


f-) \emph{Use the Box–Cox method to determine the best transformation on the 
response.}

```{R}
bc_model = MASS::boxcox(lm_, plotit=TRUE)
lambda_bc = bc_model$x[which.max(bc_model$y)]
```

The critical lambda value is at `r round(lambda_bc, 3)`. 

```{R}
train$O3_bc = (train$O3^lambda_bc-1)/lambda_bc
lm_bc_ = lm(O3_bc ~ . - O3, train)
lm_bc_summary = summary(lm_bc_)
print(lm_bc_summary)
```

The $R2$ in the original training set was 0.684. In the transformed data it is
`r round(lm_bc_summary$r.squared, 3)`.

\emph{If transformation is needed, perform the transformation and rebuild 
the model on the train data and check the model stability on the test data set.}

```{R}
predictTest <- (lambda_bc * predict(lm_bc_, test) + 1)^(1/lambda_bc)
modelTest <- cbind(predictTest, test)
testResiduals = modelTest$O3 - modelTest$predictTest
dimE = testN - k - 1

sst = sum((modelTest$O3-mean(modelTest$O3))^2)
sse = sum(testResiduals^2)
ssr = sum((modelTest$predictTest - mean(modelTest$O3))^2)

mse = sse/dimE
rse = sqrt(mse)
rmse = sqrt(sse/length(testResiduals))
rSquared = 1.0 - (sse/sst)
mae = sum(testResiduals)/length(testResiduals)

model_summary = data.frame(MSE=c(mse), RSE=rse, RMSE=c(rmse), 
                           Rsquared=c(rSquared), MAE=c(mae))
print(model_summary)
```

In the transformed training data, $R^2$ is
`r round(lm_bc_summary$r.squared, 3)`.

In the transformed test data, $R^2$ is
`r round(model_summary$Rsquared, 3)`.

```{R}
plot(testResiduals)
```

## Problem 3

\emph{Using the teengamb data under faraway library,if you are using the R 
Studio, please use the command below to download the data:}
```{r}
library(faraway)
data(teengamb, package="faraway")
```

a-) \emph{Fit a model with gamble as the response and the other variables as predictors.}

```{R}
lm_ = lm(gamble ~ ., teengamb)
lm_summary = summary(lm_)
print(lm_summary)
par(mfrow=c(2,2))
plot(lm_)
```

```{R}
par(mfrow=c(1,1))
predictors = c("status", "income", "verbal")
for (predictor in predictors) {
  plot(teengamb[,predictor], teengamb[,"gamble"], main=paste(predictor, " vs gamble"), xlab=predictor, ylab="gamble")
  abline(lm_$coefficients[["(Intercept)"]], lm_$coefficients[[predictor]])
}
```

\emph{Using the graphs, comment on the regression model and assumptions.}
From the residual and sqrt residual plots, you can see there is a difference in
variability from the low extreme, increasing at higher observed values. There is
a higher concentration of data at the low extreme.

The Q-Q Residuals plot has deviation at the ends. Points 24, 36, and 39, seem
like outliers.

From the plot of leverage, you can see that some data points have high leverage.

The model seems to be heteroscedastic.

b-) \emph{Predict the data(teengamb) amount that a male with average (given 
these data) status, income and verbal score would gamble along with an 
appropriate 95\% CI.}

```{R}
mean_male_data = data.frame(
  sex=c(0), 
  status=mean(teengamb$status), 
  income=mean(teengamb$income), 
  verbal=mean(teengamb$verbal))
mean_predictions = predict(lm_, mean_male_data, interval="confidence")
print(mean_predictions)
```

c-) \emph{Repeat the prediction for a male with maximal values (for this data) 
of status, income and verbal score.}

```{R}
max_male_data = data.frame(
  sex=c(0), 
  status=max(teengamb$status), 
  income=max(teengamb$income), 
  verbal=max(teengamb$verbal))
max_predictions = predict(lm_, max_male_data, interval="confidence")
print(max_predictions)
```

\emph{Which CI is wider and why is this result expected?}

The confidence interval for maximal values is wider. This would be expected,
because a) the variance of the residuals increases at the upper extrema b)
variability of predictions tends to increase as you move away from the center
of the data.

d-) \emph{Fit a model with $\sqrt{gamble}$ as the response but with the same 
predictors.}

```{R}
teengamb_sqrt = cbind(
  gamble_sqrt=sqrt(teengamb$gamble), 
  teengamb[,names(teengamb) != "gamble"]
)

lm_sqrt_ = lm(gamble_sqrt ~ ., teengamb_sqrt)
summary(lm_sqrt_)
```

\emph{Now predict the response and give a 95\% prediction interval for the 
individual in (a). Take care to give your answer in the original units of the 
response.}

```{R}
mean_sqrt_predictions = predict(lm_sqrt_, mean_male_data, interval="confidence")
print(mean_sqrt_predictions^2)
```

```{R}
max_sqrt_predictions = predict(lm_sqrt_, max_male_data, interval="confidence")
print(max_sqrt_predictions^2)
```

e-) \emph{Repeat the prediction for the model in (d) for a female with 
status=20, income=1, verbal = 10.}

```{R}
specific_female_data = data.frame(
  sex=1,
  status=20,
  income=1,
  verbal = 10
)
specific_sqrt_predictions = predict(lm_sqrt_, specific_female_data, interval="confidence")
print(specific_sqrt_predictions^2)
```

```{R}
par(mfrow=c(1,1))
predictors = c("status", "income", "verbal")
for (predictor in predictors) {
  plot(teengamb_sqrt[,predictor], teengamb_sqrt[,"gamble_sqrt"], main=paste(predictor, " vs gamble"), xlab=predictor, ylab="gamble")
  abline(lm_sqrt_$coefficients[["(Intercept)"]], lm_sqrt_$coefficients[[predictor]])
  points(specific_female_data[,predictor], specific_sqrt_predictions[,"fit"]^2, col="red")
}
```

\emph{Comment on the credibility of the result.}

We see from Question 1 that looking at bands can 

The predicted point seems to fall in the confidence interval of each of the 
plots. The most important predictor is income, and the prediction is taking 
place near the center of the income data.
