---
output:
  pdf_document: default
  html_document: default
---
```{r setup, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

# Linear Regression


## Libraries

The `library()` function is used to load *libraries*, or groups of  functions and data sets that are not included in the base `R` distribution. Basic functions that perform least squares linear regression and other simple analyses come standard with the base distribution,  but more exotic functions require additional libraries.

Here we load the 'faraway' and `MASS` packages, which are a very large collection of data sets and functions. We  also load the `ISLR2` package, which includes the data sets we will use.


```{r chunk1}
#install faraway package 
install.packages("faraway")
#once package is installed, you don't need to run the code in line 18
#you need to load the package every time you start the R session
library(faraway)
library(MASS)
install.packages("ISLR2")
library(ISLR2)
```


If you receive an error message when loading any of these libraries, it
likely indicates that the corresponding library has not yet been
installed on your system. Some libraries, such as `MASS`, come with `R`
and do not need to be separately installed on your computer. However,
other packages, such as `ISLR2`, must be downloaded the first time they
are used. This can be done directly from within `R`. For example, on a
Windows system,  select the `Install package` option
under the `Packages` tab.  After you select any mirror site, a
list of available packages will appear. Simply select the package you
wish to install and `R` will automatically download the
package. Alternatively, this can be done at the `R` command line
via `install.packages("ISLR2")`. This installation only needs
to be done the first time you use a package. However, the
`library()` function must be called within each `R` session.

## Simple Linear Regression

The `ISLR2` library contains the `Boston`  data set, which
records `medv` (median house value) for $506$ census tracts in Boston. We will seek to predict `medv` using $12$ predictors such as `rmvar` (average number of  rooms per house), `age` (proportion of owner-occupied units built prior to 1940) and `lstat` (percent of households with low socioeconomic status).

What is our dependent variable?

What are our independent variables?

```{r chunk2}
#head comment shows the first six observation of the data
head(Boston)
#following two lines shows the data dictionary
help(Boston)
?Boston
#str comment shows the data frame 
str(Boston)
```



We will start by using the `lm()` function to fit a simple  linear regression model, with `medv` as the response and `lstat`  as the predictor. The basic syntax is `lm(y ~ x, data)`, where `y` is the response, `x` is the predictor, and `data` is the data set in which these two variables are kept.

```{r chunk3, error=TRUE}
lm.fit = lm(medv ~ lstat,data=Boston)
```

The command causes an error because `R` does not know where to find the variables `medv` and `lstat`. The next line tells `R` that the variables are in `Boston`. If we attach `Boston`, the first line works fine because `R` now recognizes the variables.

```{r chunk4}
lm.fit <- lm(medv ~ lstat, data = Boston)
attach(Boston)
lm.fit <- lm(medv ~ lstat)
```


If we type `lm.fit`,  some basic information about the model is output. For more detailed information, we use `summary(lm.fit)`. This gives us $p$-values and standard errors for the coefficients, as well as the $R^2$ statistic and $F$-statistic for the model.


```{r chunk5}
lm.fit
summary(lm.fit)
```

We can use the `names()` function in order to find out what other pieces of information  are stored in `lm.fit`.
Although we can extract these quantities by name---e.g. `lm.fit$coefficients`---it is safer to use the extractor functions like `coef()` to access them.

```{r chunk6}
names(lm.fit)
coef(lm.fit)
```

In order to obtain a confidence interval for the coefficient estimates, we can use the `confint()` command.

```{r chunk7}
confint(lm.fit)
```

The `predict()` function can be used to produce confidence intervals and prediction intervals for the prediction of `medv` for a given value of `lstat`. We will talk about the prediction and confidence intervals later in the class.

```{r chunk8}
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))),
    interval = "confidence")
predict(lm.fit, data.frame(lstat = (c(5, 10, 15))),
    interval = "prediction")
```

For instance, the 95 \% confidence interval associated with a `lstat` value of 10 is $(24.47, 25.63)$, and the 95 \% prediction interval is $(12.828, 37.28)$.
As expected, the confidence and prediction intervals are centered around the same point (a predicted value of $25.05$ for `medv` when `lstat` equals 10), but the latter are substantially wider.

We will now plot `medv` and `lstat` along with the least squares regression line using the `plot()` and `abline()` functions.

```{r chunk9}
plot(lstat, medv)
abline(lm.fit)
```

There is some evidence for non-linearity in the relationship between `lstat` and `medv`. We will explore this issue  later.

The `abline()` function can be used to draw any line, not just the least squares regression line.
To draw a line with intercept `a` and slope `b`, we  type `abline(a, b)`. Below we experiment with some additional settings for plotting lines and points.
 The `lwd = 3` command causes the width of the regression line to be increased by a factor of 3;  this works for the `plot()` and `lines()` functions also. We can also use the `pch` option to create different plotting symbols.

```{r chunk10}
plot(lstat, medv)
abline(lm.fit, lwd = 3)
abline(lm.fit, lwd = 3, col = "red")
plot(lstat, medv, col = "red")
plot(lstat, medv, pch = 20)
plot(lstat, medv, pch = "+")
```

