---
title: "Lab03 Spring 2026 Section"
output:
  pdf_document: default
  html_document: default
date: "2026-02-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (Textbook 2.23) Refer to Copier maintenance Problem 1.19 
a. Set up the ANOVA table. 
```{r problem 2.23a}
df19 = read.table("CH01PR19.txt", 
                          header=FALSE, sep="", 
                          col.names=c("y","x"))

lmFit19 = lm(y~x, data = df19)
anova(lmFit19)

```
b. What is estimated by MSR in your ANOVA table? by MSE? Under what condition do MSR
and MSE estimate the same quantity? 

> You need to look at your ANOVA table for the values. 
MSR  = Mean SQ of X = 3.5878
MSE = Mean SQ of residulas = .3883 
They would estimate the same quantity when b1=0. Your slope has to be 0 for MSE = MSR. 


c. Conduct an F test of whether or not B1 = O. Control the a risk at .01. State the alternatives, decision rule, and conclusion. 
```{r problem 2.23c}
f = 3.5878 / .3883
print(f) 

```

> alternatives & decision rule: $H_0$: $B_1$ =0, $H_a$:$B_1$ !=0

> Conclusion: reject the null since p-value is less than our alpha (risk) at .01.

> P-value: (found in anova): .002917 < .01

> With the question in the homework, you have to take it further because you have a train and test dataset now

> Step #1 run the regression  problem above and subtract some objects to use later. This is creating your coefficient objects

> Step #2 obtain the F statistics and p-value associated with the regression

> Step #3 apply the decision rule if p-value > 0.01 (this is from this example - what does your homework say?) 

> Step #4 get the F critical

> Step #5 Confirm numerical equivalence - #Is the P-value for the F test the same as that for the t test? 


d. What is the absolute magnitude of the reduction in the variation of Y when X is introduced into the regression model? What is the relative reduction? What is the name of the latter measure? 
```{r problem 2.23d}

#you can find SSE and SSR from lmFit19 
anova(lmFit19)
SSE = 45.818 
SSR = 3.588 

#you can find SSE(R) from lmFit19_r when x= 1 (this is our reduced regression) 
lmFit19_r = lm(formula = y ~ 1, data= df19)
anova(lmFit19_r)

# SSE(R) = 49.405 

#rr = SSR / SSE
rr = 3.588/49.405 
print(rr)

```

> Relative reduction: 0.07262423

 

## (Textbook 2.30) Refer to Crime rate Problem 1.28.
a. Test whether or not there is a linear association between crime rate and percentage of high school graduates, using a t test with a = .01. State the alternatives, decision rule, and conclusion. What is the P-value of the test? 
```{r problem 2.30a}
df_crimeRate = read.table("CH01PR28.txt", 
                          header=FALSE, sep="", 
                          col.names=c("cr","ed"))
lmFit1_28 = lm(cr~ed, data=df_crimeRate)
summary(lmFit1_28)

#here we are working with a t-sided t-test

#calculcate t-value
t = -170.58 / 41.57
print(t)

#calculate critical value
#our degrees of freedom is showing that you have about 84 observations minus 2 (estimating 2 unknown parameters - slope and intercept)
critical = qt(0.995, df = 82)
print(critical)
```
> alternatives: $H_0$: $B_1$= 0 $H_a$: $B_1$ neq 0 

> t-value: -4.103
> p-value (found in summary): 9.571e-05

> decision rule: reject $H_0$ if |t| > 2.637 or reject $H_0$ if the p-value <.01

> conclusion: So we reject the null and say that our regression model is significant 


