---
title: 'CSCI E-106:Assignment 3'
output:
  pdf_document: default
  html_document:
    df_print: paged
  header-includes:
    - \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Brian Ledger

## Problem 1

Refer to the Crime rate data. A criminologist studying the relationship
between level of education-and crime rate in medium-sized U.S. counties
collected the following data for a random sample of 84 counties; X is
the percentage of individuals in the county having at least a
high-school diploma, and Y is the crime rate (crimes reported per
100,000 residents) last year. (100 points)

```{R}
set.seed(10)
crime_rate_data <- read.csv("Crime Rate.csv")
```

a-) Create train data set by selecting 70% of the data randomly and use
the remaining 30% of the data as a test data set (use set.seed(10)
before selecting the sample).

```{r}
popN = nrow(crime_rate_data)
alphaN = 7/10 * popN
idx_sample = sample(popN, size = alphaN, replace=FALSE)
sample = crime_rate_data[idx_sample,]
sampleN = nrow(sample)
test = crime_rate_data[-idx_sample,]
testN = nrow(test)

training_data = sample
test_data = test
```

**Use train data set to answer questions for part b to i**

b-) \emph{Build a regression model to predict Y on the train data set. }

```{R}
linear_model = lm(Y ~ X, training_data)
summary(linear_model)
```

\emph{Comment on your model.}

The p-value for the intercept is $0.000124 < 0.05$. This indicates that
the intercept is statistically significant.

The p-value for the slope is $0.016051 < 0.05$. This indicates that the
slope is statistically significant.

The F-statistic is $6.165 >> 1$. We would reject the hypothesis that the
slope is 0.

\emph{Are there any outliers in the data set?}

We can assess whether the errors are normally distributed by looking for
outliers in the residuals, typically, errors more than $2*\sigma$ are
considered moderate outliers.

```{R}
print(linear_model$residuals[linear_model$residuals / sigma(linear_model) > 2])
```

A single outlier is identified. The rest of the residuals tend to fit a
standard normal distribution.

\emph{Are there errors normally distributed?}

### Q-Q plot check for normally distributed residuals.

We can look at the Q-Q plot of the residuals. If the residuals are
normally distributed, they will fall into a straight line.

```{R}
plot(linear_model, which = c(2))
```

The residuals appear to deviate at the ends of the plot, suggesting they
are not normally distributed.

### Shapiro Test for normally distributed residuals.

We can run a Shapiro test to get a P-value. The null hypothesis is that
the data follow the normal distribution. The alternative hypothesis is
that the data do not follow the normal distribution.

```{R}
lm_res = linear_model$residuals
st = shapiro.test(lm_res)
st
```

The p-value for Shapiro is `r st$p.value`.

```{r}
shapiro_decision = st$p.value > 0.05
if (st$p.value > 0.05) {
  print(sprintf("%.3f > 0.05 : H_0", st$p.value))
} else {
  print(sprintf("%.3f < 0.05 : H_a", st$p.value))
}
```

### Kolmogorov-Smirnov Test for normally distributed residuals.

We can run a Kolmogorov-Smirnov test to get a P-value. The null
hypothesis is that the data follow the normal distribution. The
alternative hypothesis is that the data do not follow the normal
distribution.

```{R}
kst = ks.test(lm_res, "pnorm", mean(lm_res), sd(lm_res))
kst
```

The p-value for K.S. is `r kst$p.value`.

```{r}
if (kst$p.value > 0.05) {
  print(sprintf("%.3f > 0.05, so we fail to reject the null hypothesis.", kst$p.value))
} else {
  print(sprintf("%.3f < 0.05, so we accept the alternative hypothesis.", kst$p.value))
}
```

c-) \emph{Set up the ANOVA table for the regression model.}

Let's state the null and alternative hypotheses:

$$
\begin{aligned}
H_0: & \beta_1 = 0\\
H_a: & \beta_1 \neq 0
\end{aligned}
$$

```{R}
anova_table_constructor <- function(training_data) {
    linear_model <- lm(training_data)
    mean_y = mean(training_data[,"Y"])
    
    r_h0 = mean_y - training_data[,"Y"]
    r_h1 = linear_model$residuals
    r_r = predict(linear_model, training_data) - mean_y
    
    anova_table = data.frame(row.names = c("Regression", "Error", "Total"))
    anova_table = cbind(anova_table, "D.F."=c(
      1,
      nrow(training_data)-2, 
      nrow(training_data)-1)
    )
    anova_table = cbind(anova_table, "S.S."=c(
      sum(r_r^2),
      sum(r_h1^2), 
      sum(r_h0^2))
    )
    anova_table = cbind(anova_table, "M.S.S."=
                          c(anova_table[,"S.S."]/anova_table[,"D.F."])
    )
  return(anova_table)
}

# 

mean_y = mean(training_data[,"Y"])

r_h0 = mean_y - training_data[,"Y"]

r_h1 = linear_model$residuals

r_r = predict(linear_model, training_data) - mean_y

anova_table = data.frame(row.names = c("Regression", "Error", "Total"))

anova_table = cbind(anova_table, "D.F."=c(
  1,
  nrow(training_data)-2, 
  nrow(training_data)-1)
)

anova_table = cbind(anova_table, "S.S."=c(
  sum(r_r^2),
  sum(r_h1^2), 
  sum(r_h0^2))
)

anova_table = cbind(
  anova_table,
  "M.S.S."= c(anova_table[,"S.S."]/anova_table[,"D.F."])
)

print(anova_table)
```

```{R}
f_value = anova_table[["Regression", "M.S.S."]]/anova_table[["Error", "M.S.S."]]

p_value = 1-pf(f_value, 1, 56)
```

The F value is `r f_value`. The imputed P-value is `r p_value`.

This is confirmed by the built-in ANOVA function:

```{R}
summary(aov(linear_model))
```

d-) \emph{Carry out the test in part a by means of the T test.}

\emph{Is the P-value for the F test the same as that for the t test?}

### T-Test

The Student's t-test and the one-sided F-test (with 1 numerator degree
of freedom) are equivalent when comparing two groups or testing a single
regression coefficient.

## 

$$
F(1, df_d) = t^2(df_d)
$$

The two-tailed T-Test and the one-tailed F-Test are related thusly:

$$
F_{0.95}(1, n-2) \cong t_{.975}(n-2)^2
$$

\emph{Show the numerical equivalence of the two test statistics and decision 
rules.}

```{R}
t_squared = summary(linear_model)$coefficients["X","t value"]^2
f = anova(linear_model)["X","F value"]

print(t_squared-f)
```

The difference between $t^2$ and $f$ is negligibly small.

```{R}
critical_t_squared = qt(0.975, nrow(training_data)-2)^2
critical_f = qf(0.95, 1, nrow(training_data)-2)

print(critical_t_squared - critical_f)
```

The difference_between $(t^*)^2$ and $f^*$ is negligibly small.

e-)
\emph{By how much is the total variation in crime rate (Y) reduced when 
percentage of high school graduates (X) is introduced into the analysis?}

```{R}
print(anova_table)
```

The sum of squared residuals for the model without $X$ is
`r anova_table["Total", "S.S."]`.

The sum of squared variation for the model with $X$ is
`r anova_table["Error", "S.S."]`.

The total variation is reduced by
`r anova_table["Total", "S.S."] - anova_table["Error", "S.S."]`

\emph{Is this a relatively large or small reduction?}

```{R}
SST = anova_table["Total", "S.S."]
SSE = anova_table["Error", "S.S."]
coeff = (SSE - SST) / SST
percent_decrease = (1 - coeff) / 100
```

This constitutes a reduction of `r percent_decrease`%. This is
significant.

f-) \emph{State the full and reduced models.}

The full model is

$$
Y = \beta_0 + \beta_1 X + \varepsilon
$$ The reduced model is

$$
Y = \beta_0 + \varepsilon
$$

g-)

\begin{em} Obtain 
\begin{enumerate}
\item SSE(F), 
\item SSE(R), 
\item dfF. 
\item dfR,
\item test statistic F* for the general linear test, 
\item decision rule.
\end{enumerate}
\end{em}

These figures are equivalent to the figures in part B, and in the ANOVA
table.

```         
print(anova_table)
```

$SSE(F)$ is equivalent to $SSE$. $SSE(R)$ is equivalent to $SST$. $dfF$
is equivalent to $DFE$. $dfR$ is equivalent to $DFT$.

The test statistic $F*$ is `r qf(0.95,1, nrow(training_data)-2)`.

The decision rule is to reject the null hypothesis if $SSR/MSE < F*$,
else fail to reject.

h-) Are the test statistic F\* and the decision rule for the general
linear test numerically equivalent to those in part b?

Yes.

i-) Evaluate model performance and stability on the train data set.

Let's run it over a bunch of samples.

```{r}
RUNS = 250
run_data = list()
for (i in 1:RUNS) {
  idx_training_data = sample(popN, size = alphaN, replace=FALSE)
  training_data = crime_rate_data[idx_training_data,]
  test_data = crime_rate_data[-idx_training_data,]
  
  # Run Models
  anova_table = anova_table_constructor(training_data)
  run_data[[i]] = data.frame(
      MSR=anova_table["Regression", "M.S.S."],
      MSE=anova_table["Error", "M.S.S."]
    )
}
run_data = do.call(rbind, run_data)

msr_chi_square_test = chisq.test(run_data$MSR)
msr_chi_square_residuals = msr_chi_square_test$residuals
plot(msr_chi_square_residuals, main="MSR Chi Square Residuals")
print(msr_chi_square_test)

mse_chi_square_test = chisq.test(run_data$MSE)
mse_chi_square_residuals = mse_chi_square_test$residuals
plot(mse_chi_square_residuals, main="MSE Chi Square Residuals")
print(mse_chi_square_test)
```

You would expect the MSR and MSE of each sample, taken over multiple samples, to
form a chi-square distribution. The chi-square tests have p values on the order
of 10^{-16}, sufficient to discharge the null hypothesis.

The plots of the best-fit chi-square residuals seem evenly distributed, don't
show any non-linear effects.

**Use test data set below**

j-) Evaluate model performance and stability on the test set and compare
it to the training set. Comment on your findings? (20 points)

```{R}
RUNS = 250
run_data = list()
for (i in 1:RUNS) {
  idx_training_data = sample(popN, size = alphaN, replace=FALSE)
  training_data = crime_rate_data[idx_training_data,]
  test_data = crime_rate_data[-idx_training_data,]
  
  # Run Models
  training_anova_table = anova_table_constructor(training_data)
  test_anova_table = anova_table_constructor(test_data)
  run_data[[i]] = data.frame(
      trainingMSR=training_anova_table["Regression", "M.S.S."],
      trainingMSE=training_anova_table["Error", "M.S.S."],
      testMSR=test_anova_table["Regression", "M.S.S."],
      testMSE=test_anova_table["Error", "M.S.S."]
    )
}
run_data = do.call(rbind, run_data)

training_mse_chi_square_test = chisq.test(run_data$trainingMSE)
training_mse_chi_square_residuals = training_mse_chi_square_test$residuals
test_mse_chi_square_test = chisq.test(run_data$trainingMSE)
test_mse_chi_square_residuals = test_mse_chi_square_test$residuals
plot(training_mse_chi_square_residuals, test_mse_chi_square_residuals, main="Training vs Test MSE Chi Square Residuals")
abline(0,1)


training_msr_chi_square_test = chisq.test(run_data$trainingMSE)
training_msr_chi_square_residuals = training_msr_chi_square_test$residuals
test_msr_chi_square_test = chisq.test(run_data$trainingMSE)
test_msr_chi_square_residuals = test_msr_chi_square_test$residuals
plot(training_msr_chi_square_residuals, test_msr_chi_square_residuals, main="Training vs Test MSR Chi Square Residuals")
abline(0,1)

```

The scatterplots of MSE and MSR residuals between samples and tests are remarkably
correlated. Perhaps they even have a correlation of 1 (This could be due to an 
error in implementation).
