---
title: 'CSCI E-106:Assignment 7'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(faraway)
library(glmnet)
library(MASS)
```

## Problem 1

```{r}
data(fat,package="faraway")
str(fat)
```

\emph{Take the fat data, and use the percentage of body fat, siri, as the response and 
the other variables, except brozek and density as potential predictors.}

```{R}
cnames = names(fat)
brozek = match("brozek", cnames)
density = match("density", cnames)
fat_data = subset(fat, select = -c(brozek, density))
```

\emph{a-) Create train and test data sets. Use 70% of the data for train data 
set and use remaining data (30% of data) for test data set 
(use set.seed(1023)).}

```{R}
set.seed(1023)
fat_pop_n = nrow(fat_data)
fat_alpha_n =round(7/10*fat_pop_n)
fat_training_idx = sample(fat_pop_n, size = fat_alpha_n, replace=FALSE)
fat_training = fat_data[fat_training_idx,]
fat_testing = fat_data[-fat_training_idx,]
fat_training_n =  nrow(fat_training)
fat_testing_n = nrow(fat_testing)
print(sprintf("%i/%i=%f", fat_training_n, fat_pop_n, fat_training_n/fat_pop_n))
```

\emph{Train each model and comment on the model performance.}

a-) Linear regression with all predictors,

```{R}
fat_lm = lm(siri ~ ., fat_training)
mse = mean(fat_lm$residuals^2)
print(mse)
summary(fat_lm)
defaultSummary(data.frame(obs=fat_training$siri, pred=predict(fat_lm, fat_training)))
```

The R-squared values are very high. The model may even be over-fitted.

There are positive and negative weights in the model, which can happen when 
there is multi-collinearity.

The MSE is 1.64.

b-) Linear regression with variables selected using stepwise (both ways) 
selection criteria,

```{R}
fat_ols_step_both = olsrr::ols_step_both_adj_r2(fat_lm)
fat_ols_step_both
defaultSummary(data.frame(obs=fat_training$siri, pred=predict(fat_ols_step_both$model, fat_training)))
```

The R-squared (0.974) and Ajdusted R-squared (0.972) are almost identical to
that of the full linear model. There are 10 variables in this model, as opposed
to the 15 of the full model.

The MSE is 1.68.

c-) \emph{Ridge Regression}

```{R}
fat_training_response = fat_training$siri
fat_training_dm = model.matrix(siri ~ ., fat_training)[,-c(1)]

fat_ridge = cv.glmnet(
  fat_training_dm, 
  fat_training$siri,
  nlambda=100, 
  lambda.min.ratio=0.0001,
  type.measure="mse"
)
fat_ridge_lambda_min = fat_ridge$lambda.min
print(fat_ridge)
plot(fat_ridge)
defaultSummary(data.frame(
    obs=fat_training_response, 
    pred=c(predict(fat_ridge, s=fat_ridge_lambda_min, fat_training_dm))
  )
)

```

The R-squared of 0.974 is almost identical to the linear models.

The MSE is 1.66.

d-) Perform Lasso Regression and comment on the model performance

```{R}
fat_lasso = cv.glmnet(
  fat_training_dm, 
  fat_training_response, 
  alpha=1,
  nlambda=100, 
  lambda.min.ratio=0.0001
)
fat_lasso_lambda_min = fat_lasso$lambda.min
print(fat_lasso)
plot(fat_lasso)
defaultSummary(data.frame(
    obs=fat_training_response, 
    pred=c(predict(fat_lasso, s=fat_lasso_lambda_min, fat_training_dm))
  )
)

```

The R-squared of 0.974 is almost identical to the linear models.

The MSE is 1.67.


e-)Perform Elastic Net and comment on the model performance

```{R}
alpha_ivl = 0.01
fat_en_res = data.frame()
for (i in 1:floor(1/alpha_ivl)) {
  alpha = i*alpha_ivl
  fat_en = cv.glmnet(fat_training_dm, fat_training_response, alpha=alpha, nlambda=100,lambda.min.ratio=0.0001)
  fat_en_lambda_min = fat_en$lambda.min
  fat_en_nzero = fat_en$nzero[fat_en$lambda == fat_en_lambda_min]
  fat_en_coefficients = predict(fat_en, s=fat_en_lambda_min, type="coefficients")
  fat_en_predicted = as.matrix(data.frame(1, fat_training_dm)) %*% fat_en_coefficients
  fat_training_sst = sum((fat_training_response - mean(fat_training_response))^2)
  fat_en_sse = sum((fat_training_response - fat_en_predicted)^2)
  fat_en_r_squared = 1 - (fat_en_sse/fat_training_sst)
  fat_en_res = rbind(
    fat_en_res, 
    data.frame(alpha=alpha, nzero=fat_en_nzero, r_squared=fat_en_r_squared)
  )
}

plot(fat_en_res$alpha, fat_en_res$r_squared, main="ElasticNet", xlab="alpha", ylab="R^2")
plot(fat_en_res$alpha, fat_en_res$nzero, main="ElasticNet", xlab="alpha", ylab="nzero")
  
best_r_squared = max(fat_en_res$r_squared)
best_alpha = fat_en_res$alpha[fat_en_res$r_squared==best_r_squared]
fat_en = cv.glmnet(fat_training_dm, fat_training_response, alpha=best_alpha, nlambda=100,lambda.min.ratio=0.0001)
fat_en_lambda_min = fat_en$lambda.min

print(fat_en)
plot(fat_en,)
defaultSummary(data.frame(
    obs=fat_training_response, 
    pred=c(predict(fat_en, s=fat_en_lambda_min, fat_training_dm))
  )
)

```

The Elastic Net regression converges to the lasso.

The MSE is 1.66.

**Use the test data sets to answer questions below**

f-) Test the model's performances for part (a) to part (e) above and comment on 
model stability and performance. Which model would you choose and why?

```{R}
fat_testing_dm = model.matrix(siri ~ ., fat_testing)[,-c(1)]

# Linear Model
defaultSummary(data.frame(obs=fat_testing$siri, pred=predict(fat_lm, fat_testing)))

# Stepwise Two Way Ajusted R-Squared
defaultSummary(data.frame(obs=fat_testing$siri, pred=predict(fat_ols_step_both$model, fat_testing)))

# Ridge
defaultSummary(data.frame(obs=fat_testing$siri, pred=c(predict(fat_ridge, s=fat_ridge_lambda_min, fat_testing_dm))))

#Lasso
defaultSummary(data.frame(obs=fat_testing$siri, pred=c(predict(fat_lasso, s=fat_lasso_lambda_min, fat_testing_dm))))

# ElasticNet
defaultSummary(data.frame(obs=fat_testing$siri, pred=c(predict(fat_en, s=fat_en_lambda_min, fat_testing_dm))))


```
The R-squared are lower in test, which may suggest overfitting. The Lasso has
the lowest RMSE, and MAE, and highest R-squared. I would use Lasso.

## Problem 2

Use the fat data set in previous example by using the percentage of body fat, 
siri, as the response and the other variables, except brozek and density as 
potential predictors.

a-)Fit the same model in question 1-a but now using Huber's robust method. 
Comment on any substantial differences between this model and the least squares 
fit.

```{R}
fat_huber = rlm(siri ~ ., fat_training)
fat_hampel = rlm(siri ~ ., fat_training, psi=psi.hampel)
fat_bisquare = rlm(siri ~ ., fat_training, psi=psi.bisquare)
fat_rlm = data.frame(
  idx = c(1:nrow(fat_training)),
  huber = fat_huber$w,
  hampel = fat_hampel$w,
  bisquare = fat_bisquare$w
)
fat_rlm = fat_rlm[order(fat_rlm$huber),]
head(fat_rlm)
```

b-)Identify which two cases have the lowest weights in the Huber fit. What is 
unusual about these two points?

```{R}
fat_outlier_idxs = fat_rlm[1:6, "idx"]
fat_outlier_residuals = fat_huber$residuals[fat_outlier_idxs]


plot(
  fat_huber$residuals,
  pch = 19,
  cex=0.25,
  xlab="Residuals",
  ylab = "",
)

points(fat_outlier_idxs[1:2], fat_outlier_residuals[1:2], col="red")
points(fat_outlier_idxs[3:4], fat_outlier_residuals[3:4], col="orange")
points(fat_outlier_idxs[5:6], fat_outlier_residuals[5:6], col="violet")
```

These points have the most extreme residual values.

```{R}
plot(
  fat_training$weight, 
  fat_training$siri, 
  pch = 19, 
  cex=0.25,
  main="Weight vs. Siri",
  xlab="Weight",
  ylab="Siri"
  )
points(fat_training[fat_outlier_idxs[1:2],"weight"], fat_training[fat_outlier_idxs[1:2],"siri"],  col="red")
points(fat_training[fat_outlier_idxs[3:4],"weight"], fat_training[fat_outlier_idxs[3:4],"siri"], col="orange")
points(fat_training[fat_outlier_idxs[5:6],"weight"], fat_training[fat_outlier_idxs[5:6],"siri"], col="violet")
```

The most extreme huber weight corresponds to a row of maximum weight over all of 
the data.

c-)Plot weight (of the man) against height. Identify the two outlying cases. Are 
these the same as those identified in the previous question? Discuss.

```{R}
plot(
  fat_training$weight, 
  fat_training$height, 
  pch = 19, 
  cex=0.25,
  main="Weight vs. Height",
  xlab="Weight",
  ylab="Height"
)
points(
  fat_training[fat_outlier_idxs[1:2],"weight"], 
  fat_training[fat_outlier_idxs[1:2],"height"],  
  col="red"
)
```

The indices of the outliers point to at least one high leverage point in the
weight vs height chart.