---
title: "Variable Selection R Codes Examples"
author: "Hakan Gogtas"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Manual Backward Variable Selection

How backward elimination works\newline
The process follows a clear, ordered loop:\newline
- Fit the full model with all predictors.\newline
- Identify the predictor with the largest p‑value.\newline
- If that p‑value is > 0.05, remove that predictor.
- Refit the model without it.\newline
- Repeat until all remaining predictors have p < 0.05.\newline
This produces a final model where every included variable is individually significant conditional on the others.\newline

```{r}
library(olsrr)
library(datasets)

f <- lm(mpg ~., data = mtcars)
summary(f)
#cyl has the highest p value of 0.9161
f1<-update(f,.~.- cyl)
summary(f1)
#vs has the highest p value of 0.8433
f1<-update(f1,.~.- vs)
summary(f1)
#carb has the highest p value of  0.7470
f1<-update(f1,.~.- carb)
summary(f1)
#gear has the highest p value of  0.61964
f1<-update(f1,.~.- gear)
summary(f1)
#drat has the highest p value of  0.46240
f1<-update(f1,.~.- drat)
summary(f1)
#disp has the highest p value of  0.29897 
f1<-update(f1,.~.- disp)
summary(f1)
#hp has the highest p value of  0.22309 
f1<-update(f1,.~.- hp)
summary(f1)
```

##Best Subset Regression

Select the subset of predictors that do the best at meeting some well-defined objective criterion,  such as having the largest R2 value or the smallest MSE, Mallow’s Cp or AIC.

```{r}
library(olsrr)
library(datasets)
exp <- lm(mpg ~wt+ qsec+ am, data = mtcars)
k <- ols_step_all_possible(exp)
k
plot(k)
#lets get the best model for each subset
exp1 <- lm(mpg ~., data = mtcars)
k1<-ols_step_best_subset(exp1)
k1
plot(k1)
```


The plot method shows the panel of fit criteria for best subset regression methods.


##Stepwise Forward Regression

Build regression model from a set of candidate predictor variables by entering predictors based on p values, in a stepwise manner until there is no variable left to enter any more. The model should include all the candidate predictor variables. If details is set to TRUE, each step is displayed.


```{r}
# stepwise forward regression
exp2 <- lm(y ~ ., data = surgical)
k2<-ols_step_forward_p(exp2)
plot(k2)
# stepwise forward regression
ols_step_forward_p(exp2, details = TRUE)
```

##Stepwise Backward Regression

Build regression model from a set of candidate predictor variables by removing predictors based on p values, in a stepwise manner until there is no variable left to remove any more. The model should include all  the candidate predictor variables. If details is set to TRUE, each step is displayed.


```{r}

# stepwise backward regression
k4<-ols_step_backward_p(exp2)
k4
plot(k4)
#get the model
k4$model
formula(k4$model)
#Detailed Output
# stepwise backward regression
ols_step_backward_p(exp2, details = TRUE)
```

##Stepwise Regression
Build regression model from a set of candidate predictor variables by entering and removing predictors based on p values, in a stepwise manner until there is no variable left to enter or remove any more. The model should include all the candidate predictor variables. If details is set to TRUE, each step is displayed.

```{r}

# stepwise regression

k5<-ols_step_both_p(exp2)
k5
plot(k5)
formula(k5$model)
#ensuring that all variables are significant at 95% level
k5.1<-ols_step_both_p(exp2,p_enter=0.1,p_remove=0.05,details=TRUE)
k5.1
formula(k5.1$model)
#Detailed Output
ols_step_both_p(exp2, details = TRUE)
```

##Stepwise AIC Forward Regression
Build regression model from a set of candidate predictor variables by entering predictors based on Akaike Information Criteria, 
in a stepwise manner until there is no variable left to enter any more. The model should include all the candidate predictor 
variables. If details is set to TRUE, each step is displayed.

```{r}
# stepwise aic forward regression
k6<-ols_step_forward_aic(exp2)
k6
plot(k6)
ols_step_forward_aic(exp2, details = TRUE)
```

## Stepwise AIC Backward Regression
Build regression model from a set of candidate predictor variables by removing predictors based on Akaike Information Criteria, in a stepwise manner until there is no variable left to remove any more. The model should include all the candidate predictor variables. If details is set to TRUE, each step is displayed.


```{r}
k7 <- ols_step_backward_aic(exp2)
k7
plot(k7)
ols_step_backward_aic(exp2, details = TRUE)
```


## Stepwise AIC Regression
Build regression model from a set of candidate predictor variables by entering and removing predictors based on Akaike Information Criteria, in a stepwise manner until there is no variable left to enter or remove any more. The model should include all the candidate predictor variables. 
If details is set to TRUE, each step is displayed.


```{r}
k8<-ols_step_both_aic(exp2)
k8
plot(k8)
ols_step_both_aic(exp2, details = TRUE)
```

##Example of using leaps library

We illustrate backward elimination on some data on the 50 states from the 1970s. The data were collected from U.S. Bureau of the Census. We will take life expectancy as the response and the remaining variables as predictors:


```{r}
data(state)
state.x77[1:10,]
state<-data.frame(state.x77)
require(leaps)
b<-regsubsets(Life.Exp~.,data=state)
rs<-summary(b)
rs$which

AIC<-50*log(rs$rss/50)+(2:8)*2
plot(AIC~I(1:7),ylab="AIC",xlab="Number of Predictors")

plot(2:8,rs$adjr2,xlab="Number of Predictors",ylab="Adjusted R Square")
which.max(rs$adjr2)


plot(2:8,rs$cp,xlab="Number of Predictors",ylab="Cp")
abline(0,1)

st1<-lm(Life.Exp~.,data=state)
res1<-ols_step_best_subset(st1)
res1
plot(res1)
```

The competition is between the four-parameter, three-predictor, model including frost, high school graduation and murder and the model also including population. 

Both models are on or below the Cp = p line, indicating good fits. 
The choice is between the smaller model and the larger model, which fits a little better. 

Some even larger models fit in the sense that they are on or below the Cp = p line, but we would not opt for these in the presence of smaller models that fit.



