---
title: 'CSCI E-106:Assignment 6'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(lmtest))
suppressPackageStartupMessages(library(faraway))
library(MASS)
library(car)
library(caret)
library('olsrr')
set.seed(1023) 
```

## Problem 1

```{r}
library(faraway)
data(divusa,package="faraway")
str(divusa)
```

\textbf{a-) Fit a regression model with divorce as the response and unemployed, 
femlab, marriage, birth and military as predictors.}

```{R}
d_data = divusa[
  ,c("divorce", "unemployed", "femlab", "marriage", "birth", "military")
]
d_model = lm(divorce ~ ., d_data)
summary(d_model)
```
\textbf{b-) For the same model, compute the VIFs. Is there evidence that 
collinearity causes some predictors not to be significant? Explain.}


```{R}
vif(d_model)
print("R_i^2")
print(1 - 1/vif(d_model))
```

Typically, VIF between 1 and 5 indicates moderate correlation. The highest VIF
is for `femlab`.

```{R}
d_femlab_model = lm(femlab ~ . -divorce, d_data)
summary(d_femlab_model)
plot(d_data$femlab, d_data$unemployed)
```


\textbf{c-) Does the removal of insignificant predictors from the model reduce 
the collinearity? Investigate.}


```{R}
rms = function(xs) { sqrt(mean(xs^2)) }
cat(sprintf("Baseline: %f\n", rms(vif(d_model))))
for (factor in c("femlab", "unemployed", "marriage", "birth", "military")) {
  expr = sprintf("divorce ~ . - %s", factor)
  cat(sprintf("%s: %f\n", expr, rms(vif(lm(as.formula(expr), d_data)))))
}

```
The RMS of the VIF factors decrease when we remove any of `femlab`, 
`unemployed`, `marriage`, or `birth`.


\textbf{d-) Visually using the appropriate graphs, comment on the regression 
model assumptions.}


```{R}
cor(d_data)
pairs(d_data[,-1])
plot(d_model$residuals[order(d_data$divorce)])
```

There are significant non-linear effects between almost all of the factors.

There appears to be a non-linear pattern in the residuals. There appears to be
non-uniform variance in the residuals.

## Problem 2

\textbf{Refer to the Machine speed Data, showing the number of defective items 
produced by a machine (Y) and the speed setting of the machine (X). Answer the 
questions below.}

```{R}
m_data = read.csv("Machine Speed.csv")
m_data = m_data[order(m_data$X), ]
tail(m_data)
```


\textbf{a-) Fit a linear regression function by ordinary least squares, obtain 
the residuals, and plot the residuals against X. Plot the squared residuals 
against X.}

```{R}
m_lm = lm(Y ~ X, m_data)
summary(m_lm)
```

```{R}
plot(m_lm$residuals)
plot(m_lm$residuals^2)
```

\textbf{What do the plots suggest?}

The variance appears to increase with $X$.

b-) Conduct the Breusch-Pagan test for constancy of the error variance,  use 
$\alpha$ = .10. State the alternatives, decision rule, and conclusion.

Null Hypothesis: The error variance is homoscedastic.

Alternative Hypothesis: The error variance is heteroscedastic.

```{R}
bptest(m_lm)
```

At $\alpha = 0.10$, we have $0.07 < 0.10$, so we cannot discharge the null
hypothesis.

\textbf{c-)  Calculate the weights for weighted least square estimation and 
refit the model.}

```{R}
residuals_x = data.frame(residuals = abs(m_lm$residuals), X= m_data$X)
var_est = predict(lm(residuals ~ X, residuals_x), new_data = m_data$X)
weights = 1 / var_est
m_lm_weighted = lm(Y~X, m_data, weights=weights)
summary(m_lm_weighted)
```

\textbf{d-)  Are the weighted least squares estimates of $\beta_0$ and $\beta_1$ 
similar to the ones obtained with ordinary least squares in part (a)?}

```{R}
print(m_lm$coefficients)
print(m_lm_weighted$coefficients)
```

The weights are similar in order and direction. They differ slightly.

\textbf{e-) Compare the estimated standard deviations of the weighted least 
squares estimates b$\beta_{w0}$ and $\beta_{w1}$in part (c) with those for the 
ordinary least squares estimates in part (a). What do you find?}

```{R}
print(summary(m_lm)$coefficients[, "Std. Error"])
print(summary(m_lm_weighted)$coefficients[, "Std. Error"])
```

The standard errors for the weighted model are slightly lower.

\textbf{f-) Iterate the steps in parts (c) one more time. Is there a substantial 
change in the estimated regression coefficients? If so, what should you do?}

```{R}
residuals_x = data.frame(residuals = abs(m_lm_weighted$residuals), X= m_data$X)
var_est = predict(lm(residuals ~ X, residuals_x), new_data = m_data$X)
weights = 1 / var_est
m_lm_weighted2 = lm(Y~X, m_data, weights=weights)
summary(m_lm_weighted2)
```

The estimates, Multiple R-squared, and Std. Error do not change significantly.

If there was a marginal improvement over the last iteration, you would probably
want to continue iterating until the improvement becomes insignificant.

## Problem 3

\textbf{Refer to the CDI data set used in Assignment 5. For this analysis, use 
all variables except Identification Number, County, and State. In addition, 
include  the dummy variables you created for Geographic Region in Question 1(a) 
of Assignment 5, and remove the original Geographic Region variable. The 
dependent variable is the number of active physicians (Y), which should be 
regressed on all remaining variables, including the geographic‑region dummy 
variables.}

```{R}
cdi_data_raw = read.csv("CDI Data.csv")
ignore = c("Identification.number", "County", "State")
cdi_data = subset(cdi_data_raw, select = -match(ignore, names(cdi_data_raw)))
cdi_data$Geographic.region = factor(cdi_data$Geographic.region)
tail(cdi_data)
```

\textbf{a-) Use a stepwise variable selection method to identify the best 
regression model. Then evaluate the model assumptions using appropriate 
diagnostic plots and comment on whether the assumptions appear to be satisfied.}

```{R}
cdi_model = lm(Number.of.active.physicians ~ ., cdi_data)
all_possible_cdi_models = ols_step_all_possible(cdi_model)
all_possible_cdi_models
```

The simple regression against $Number.of.hospital.beds$ gives the highest
adjusted R-Squared of 0.903.

```{R}
cdi_bestmodel = lm(Number.of.active.physicians ~ Number.of.hospital.beds, cdi_data)
summary(cdi_bestmodel)
plot(cdi_bestmodel$residuals[order(cdi_data$Number.of.active.physicians)])
```
The error variance is clearly heteroscedastic.

\textbf{b-) The CDI data set contains 440 observations. Randomly divide the data 
into five folds, each containing 88 observations, using set.seed(1023) to ensure 
reproducibility. Perform 5‑fold cross‑validation to assess the stability of your 
selected model in part a. For each fold, report the following metrics for both 
the training and testing sets:}
\begin{itemize}
\item \textbf{Root Mean Squared Error (RMSE)}
\item \textbf{R‑squared (R²)}
\item \textbf{Mean Absolute Error (MAE)}
\end{itemize}

```{R}
cdi_train_kfold <- trainControl(method = "cv", number = 5, savePredictions = "all")
cdi_lm_cv <- train(Number.of.active.physicians ~ Number.of.hospital.beds, 
                   cdi_data, method = "lm", trControl = cdi_train_kfold)
print(cdi_lm_cv$resample)
```

\textbf{Summarize the results across all five folds and comment on the model’s 
stability and predictive performance.}

```{R}
colMeans(cdi_lm_cv$resample[,c("RMSE", "Rsquared", "MAE")])
```

The range of Rsquared is ~17%. The least Rsquared is ~0.81. This model explains
most of the variability in the number of active physicians.
