---
title: "MLR Part1"
output: pdf_document
---
Now let's look at an example concerning the number of species found on the various Galápagos Islands. There are 30 cases (Islands) and seven variables in the dataset. We start by reading the data into R and examining it:


```{r}
load(url("https://raw.githubusercontent.com/cran/faraway/master/data/gala.rda"))
head(gala[,-2])
```

The variables are Species — the number of species found on the island, Area — the area of the island (km2), Elevation — the highest elevation of the island (m), Nearest — the distance from the nearest island (km), Scruz — the distance from Santa Cruz Island (km), Adjacent — the area of the adjacent island (km2). We have omitted the second column (which has the number of endemic species) because we shall not use this alternative response variable in this analysis.


```{r}
lmod<-lm(Species~Area+Elevation+Nearest+Scruz+Adjacent,data=gala)
summary(lmod)
```
Lets extract X-Matrix

```{r}
X<-model.matrix(lmod)
```

Lets extract Y-Matrix

```{r}
Y<-gala$Species

```

Lets construct Hat matrix $(X^{T}X)^{-1}X^{T}$
```{r}
H<-solve(t(X)%*%(X))%*%t(X)
```

Lets calculate $\beta$

```{r}
Beta<-H%*%Y
Beta
```


We can extract the regression quantities we need from the model object. Commonly used are residuals (), fitted (), df.residual () which gives the degrees of freedom, deviance () which gives the RSS and coef () which gives the $\hat{\beta}$. You can also extract other needed quantities by examining the model object and its summary:

We can estimate $\sigma$ using the formula in the text above or extract it from the summary object:
```{r}
names(lmod)
lmodsum<-summary(lmod)
names(lmodsum)
sqrt(deviance(lmod)/df.residual(lmod))
lmodsum$sigma
```


We can also extract $(X^{T}X)^{-1}$ and use it to compute the standard errors for the coefficients. (diag () returns the diagonal of a matrix):

```{r}
xtxi<-lmodsum$cov.unscaled
sqrt(diag(xtxi))*lmodsum$sigma
summary(lmod)
lmodsum$coef[,2]
```

## Creating Dummy Variables

Bringing data into R 

```{r}
IID <- read.csv("/cloud/project/Data/Insurance Inovation Data.csv")
```

## Using Libaries

it will create  dummy variables for categorical variables. In this case, it will create two dummy variables but we need only one, you need to manually drop it or don't use it

```{r}
library('fastDummies')
IID.Dat<-dummy_cols(IID, select_columns = 'Firm.Type')
IID.Dat[1:3,]
f<-lm(Y~X1+Firm.Type_Stock,data=IID.Dat)
summary(f)
```
## using factor command

Factor command will automatically create a dummy variable. For variable selection, it is not recommended.

```{r}
f<-lm(Y~X1+factor(Firm.Type),data=IID)
summary(f)
```


## manually coding

This is the recommended approach
```{r}
IID$X.Stock=I(IID$Firm.Type=="Stock")*1
IID$X.Stock
#alternatively
IID$X.Mutual=I(IID$Firm.Type=="Mutual")*1
IID$X.Mutual

f<-lm(Y~X1+X.Stock,data=IID)
summary(f)

# Create side-by-side boxplots
boxplot(IID$Y ~ IID$Firm.Type,
        col = c("skyblue", "lightgreen"),
        main = "Side-by-Side Boxplots",
        xlab = "Firm Type",
        ylab = "Revenue")

```

Box plots overlap. Revenues for Mutual is ranged from 0 to 30. While, for stock, 10 to 40.


## Regression Lines for Mutul and Stock Companies

```{r}
Y=IID$Y
X1<-IID$X1
X2<-IID$X2
cX2<-as.factor(IID$X2)  
fit<-lm(Y~X1 +cX2)
coef <- coefficients(fit)
plot( Y ~ X1, pch=(15+X2),col = as.character(2+X2))  
abline(coef["(Intercept)"] + coef["cX21"], coef["X1"], col = 3)
abline(coef["(Intercept)"] , coef["X1"], col = 2)
legend(35,10,c("Stock firm","Mutual firm"),col=c(3,2), lwd=1,lty=c(1,1), pch=c(16,15))
```


## the contrast matrix for categorical variable with two levels

```{r}
contr.treatment(2)
IID$Firm.Type.f<-factor(IID$Firm.Type)
contrasts(IID$Firm.Type.f) = contr.treatment(2)
f<-lm(Y~X1+Firm.Type.f,data=IID)
summary(f)
```

## Intaraction Terms
```{r}
#1st way interaction term
f<-lm(Y~X1+X2+X1*X2,data=IID)
summary(f)

#2nd way interaction term
f<-lm(Y~X1+X2+X1:X2,data=IID)
summary(f)

#3rd way interaction term
IID$X3=IID$X1*IID$X2
f<-lm(Y~X1+X2+X3,data=IID)
summary(f)

#4th way
f1<-lm(Y~.^2,data=IID[,1:3])
summary(f1)
```

## Example
```{r}
DwaineStudio <- read.csv("/cloud/project/Data/DwaineStudio.csv")
round(cor(DwaineStudio),2)
plot(DwaineStudio)
summary(DwaineStudio)
f.dw<-lm(Y~X1+X2,data=DwaineStudio )
summary(f.dw)
par(mfrow=c(2,2))
plot(f)
anova(f.dw)
```


#################
#Box Cox Transformation
#################

$\lambda$ is very close to zero. To easy up the computation, we will use the log transformation.
```{r}
library(MASS)
par(mfrow=c(1,1))
boxcox(f.dw,lambda=seq(-5,5,by=.1))
#getting the value of lambda
b=boxcox(f.dw,lambda=seq(-5,5,by=.1))
lambda<-b$x[which.max(b$y)]
lambda
#doing the log transformation
f1.dw<-lm(log(Y)~X1+X2,data=DwaineStudio )
summary(f1.dw)

#checking visually reg model assumptions
par(mfrow=c(2,2))
plot(f1.dw)
```


## Correlation Test for Normality

$H_o$: Errors are normally distributed\newline
$H_a$: Errors are Not normally distributed
```{r}
shapiro.test(f1.dw$residuals)
```

Accept Ho, it is normally distributed

## Brown-Forsythe Test

$H_o$: Error variances are constant\newline
$H_a$: Error variances are not constant

We will use X1 to split the data in two parts
```{r}
pop.dw<-data.frame(cbind(DwaineStudio$Y,DwaineStudio$X1, DwaineStudio$X2,ei=f1.dw$residuals))
library(car)
#Create the group variable based on medium X1 value (Low Xs =0 vs High Xs =1)
group=I(pop.dw[,2]< median(DwaineStudio$X1))*1
#using built in function by inputting errors ei=pop.dw[,4] and group variable as inputs
leveneTest(pop.dw[,4], group, center=median)
```

P Value is greater than 0.05. Accept $H_o$,Error variances are constant.

## Breusch-Pagan Test

$H_o$: Error variances are constant\newline
$H_a$: Error variances are not constant

```{r}
library(lmtest)
bptest(log(Y)~X1+X2,data=DwaineStudio)
```

P Value is greater than 0.05. Accept $H_o$,Error variances are constant.

## can we drop X2?

$H_o$: $\beta_{2}=0$\newline
$H_a$: $\beta_{2} \neq 0$\newline

```{r}
#model without X2
f2.dw<-lm(log(Y)~X1,data=DwaineStudio)
summary(f2.dw)
anova(f2.dw,f1.dw)
```

p value is 0.0213 <0.05, Reject $H_o$. We CAN NOT drop $X_2$



