---
title: "MLR Part II"
output:
  pdf_document: default
  html_document: default
---
## Multiple Linear Regression

The `Boston` data set contains 12 variables, and so it would be cumbersome to have to type all of these in order to perform a regression using all of the predictors.
Instead, we can use the following short-hand:

```{r}
library(MASS)
library(ISLR2)
lm.fit <- lm(medv ~ ., data = Boston)
summary(lm.fit)
```

We can access the individual components of a summary object by name
(type `?summary.lm` to see what is available). Hence
`summary(lm.fit)$r.sq` gives us the $R^2$, and
`summary(lm.fit)$sigma` gives us the RSE. 

## General F test, Can we drop Age and Indus from the model?

$H_o$: $\beta_{Age}=\beta_{Indus}=0$\newline
$H_a$: One of the $\beta$  is significant

Reduced model without Age and Indus

```{r}
lm.red<- lm(medv ~ crim+zn+chas+nox+rm+dis+rad+tax+ptratio+lstat,  data = Boston)
anova(lm.red,lm.fit)
```
Here Model 1 represents the reduced model, while Model 2 corresponds to full model.
The `anova()` function performs a hypothesis test comparing the two models. P value is 0.94, we will accept $H_o$, we can drop Age and Indus.

## Dropping variables

What if we would like to perform a regression using all of the significant variables?  
For example, in the above regression output,  `age` has a high $p$-value. So we may wish to run a regression excluding this predictor.

The following syntax results in a regression using all predictors except `age`.

```{r}
lm.fit1 <- lm(medv ~ . - age, data = Boston)
summary(lm.fit1)
```

Alternatively, the `update()` function can be used.

```{r}
lm.fit1 <- update(lm.fit, ~ . - indus )
summary(lm.fit1)
```

Lets also get rid of indus variable that is also not significant

```{r}
lm.fit2 <- update(lm.fit1, ~ . - age)
summary(lm.fit2)
#we can drop them at the same time

lm.fit2 <- update(lm.fit, ~ . - age - indus)
summary(lm.fit2)
```

All variables are significant and $R^2$ is %74.

## regression model assumptions

```{r}
par(mfrow=c(2,2))
plot(lm.fit2)
par(mfrow=c(1,1))
library(MASS)
par(mfrow=c(1,1))
boxcox(lm.fit2,lambda=seq(-5,5,by=.1))
#getting the value of lambda
b=boxcox(lm.fit2,lambda=seq(-5,5,by=.1))
lambda<-b$x[which.max(b$y)]
lambda

lm.fit3<-lm(log(medv) ~ crim+zn+chas+nox+rm+dis+rad+tax+ptratio+lstat,  data = Boston)
summary(lm.fit3)
par(mfrow=c(2,2))
plot(lm.fit3)
```

Box-Cox suggest log transformation distribution. QQ plot shows outliers. Graphs and Breusch Pagan test shows that the variances are unequal. The normality tests indicate that data is not normal.

##Breusch Pagan Test

$H_o$: Error variances are constant\newline
$H_a$: Error variances are not constant

```{r}
library(lmtest)
bptest(lm.fit2)
#on transformed model
bptest(lm.fit3)
```

## Correlation Test for Normality

$H_o$: Errors are normally distributed\newline
$H_a$: Errors are Not normally distributed
```{r}
shapiro.test(lm.fit2$residuals)
```

## Transforming the data

```{r}
lm.fit3<- lm(log(medv) ~ crim + zn + chas + nox + rm + dis + rad + tax + ptratio +  lstat, data = Boston)
summary(lm.fit3)
par(mfrow=c(2,2))
plot(lm.fit3)
bptest(lm.fit3)
```

All variables are significant and $R^2$ is %79. QQ plot shows heavy tails with no outliers. Graphs and Breusch Pagan test shows that the variances are unequal. 